---
title: "Component interpretation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Component interpretation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 4,
  fig.height = 4,
  dpi = 120
)
library(landscapeR)
```

This workflow interprets one molecular layer measured on canonical biological
samples. Registered strategies cover independent cross-sectional samples and
independent destructive-sampling time courses. Repeated-subject time courses
require their own subject-aware strategy and are not silently treated as
independent. Every result remains exploratory.

The workflow separates five actions:

1. `associate_metadata()` stores raw, adjusted, and resampled evidence.
2. `plot()` renders the stored decision surface without changing it.
3. `propose_component()` ranks only the predeclared effect and may compute a
   search-aware permutation null.
4. `assess_component_identifiability()` repeats the complete search under
   design-preserving resampling and aligns every replicate to the frozen
   discovery basis.
5. `confirm_component()` records a separate analyst decision and rationale.

## Synthetic worked example

The example demonstrates software behavior with synthetic coordinates. It is
not biological evidence and does not calibrate a universal decision rule.

```{r data}
primary <- sprintf("sample_%02d", 1:8)
assay_ids <- sprintf("rna_%02d", 1:8)
std <- StateTransitionData(
  experiments = list(
    rna = SummarizedExperiment::SummarizedExperiment(
      assays = list(logcounts = matrix(
        seq_len(32L),
        nrow = 4L,
        dimnames = list(sprintf("gene_%02d", 1:4), assay_ids)
      ))
    )
  ),
  colData = S4Vectors::DataFrame(
    condition = factor(
      rep(c("control", "treatment"), each = 4L),
      levels = c("control", "treatment")
    ),
    severity = c(1, 2, 2, 4, 5, 6, 7, 8),
    state = ordered(
      c("early", "early", "middle", "middle",
        "late", "late", "late", "end"),
      levels = c("early", "middle", "late", "end")
    ),
    subtype = factor(rep(c("A", "B", "C", "D"), each = 2L)),
    batch = rep(c("run_1", "run_2"), 4L),
    mouse_id = sprintf("mouse_%02d", 1:8),
    row.names = primary
  ),
  sampleMap = S4Vectors::DataFrame(
    assay = factor(rep("rna", 8L), levels = "rna"),
    primary = primary,
    colname = assay_ids
  )
)
std <- declare_sampling_design(std, cross_sectional())
std_metadata <- S4Vectors::metadata(std)
std_metadata$stage1 <- DecompositionResult(
  V_star = c(1, 0, 0, 0),
  sigma = 1,
  coords = list(1:8),
  V_k = diag(4)[, 1:2, drop = FALSE],
  sigma_k = matrix(c(2, 1), nrow = 1L),
  coords_k = list(cbind(
    PC1 = c(4, 3, 2, 1, 1, 2, 3, 4),
    PC2 = 1:8
  )),
  k = 2L
)
S4Vectors::metadata(std) <- std_metadata
```

## Declare intent, then build the atlas

The specification owns the target, direction, nuisance fields, and run
identity. Numerical spacing is explicit here because `severity` is declared as
continuous. Ordered factor labels instead use their declared level order and
never imply unequal spacing.

```{r atlas}
specification <- analysis_specification(
  id = "synthetic-severity",
  target_field = "severity",
  target_type = "continuous",
  continuous_direction = "increasing",
  nuisance_fields = "batch"
)

atlas <- associate_metadata(
  std,
  specification = specification,
  non_analytical_fields = "mouse_id",
  dataset_id = "synthetic-cross-sectional-control",
  n_resamples = 49L,
  seed = 7001L
)
atlas_associations(atlas)
atlas_exclusions(atlas)
atlas_evidence_contract(atlas)
plot(atlas)
```

The table keeps unadjusted and adjusted rows separate. Each row records its
complete-cohort and design digests, missingness, tied score and target counts,
bootstrap interval, failed-resample count, raw p-value, and Holm-adjusted
q-value. The correction family is the declared component set within each
metadata field and evidence variant. Bootstrap draws preserve discrete target and nuisance-cell counts and
resample complete independent biological units rather than features or
technical replicates.

Raw points, a black monotone-constrained fit, and a red flexible smoother
expose the predeclared monotone relationship. A reversal in level-wise medians adds a
`possible-nonmonotone-association` warning. The warning and smoother are
descriptive and cannot rerank components.

The evidence summary is the inspection-friendly view of the package-owned
internal interpretation contract behind the atlas. It records normalized row
counts, exact available-case cohort membership, and deterministic table and
cohort-membership digests. Atlas validity checks those values before proposal,
permutation, plotting, or serialization consumers can use the evidence.
Registered association strategies still calculate the scientific estimands;
they do not assemble or validate the surrounding evidence.

Cross-sectional and independent destructive time-course analyses now pass
through the same normalized execution kernel. This is a software consistency
boundary, not a shared statistical model. The two designs retain their own
cohorts, estimands, exchangeability rules, refits, diagnostics, and permutation
schemes. The shared kernel checks strategy contracts, traverses the declared
components, accounts for every requested resample, applies the declared
multiplicity correction, and constructs the typed atlas. Repeated-subject
analysis remains subject-aware and is never routed through an independence
assumption.

### One resampling contract, three scientific adapters

Bootstrap and permutation work share one internal policy for deterministic
seeds, immutable plan digests, requested and completed counts, failure
accounting, and typed unavailable outcomes. The requested count is fixed before
refitting. A failed eligible refit remains in that denominator and is never
silently replaced with another draw.

Independent and repeated time-course bootstraps use the package's shared
`future.apply` execution seam. Set the future backend outside landscapeR; the
package never changes `future::plan()`. A workflow already running inside an
outer future can pass `sequential_internal = TRUE` to `associate_metadata()`
to avoid nested worker multiplication. `future_scheduling` is optional and has
no package-defined default. The stored execution record includes the
`inspect`, `standard`, or `evidence` compute tier, the run seed, deterministic
task-specific L'Ecuyer-CMRG streams, all requested tasks, and stable failure
codes. These fields make results reproducible across worker counts and
scheduling without changing the biological resampling unit.

The same execution contract governs complete-search permutations: labels or
reduced-model residuals are rearranged according to the declared design, every
requested refit remains in the denominator, and failed refits are retained as
typed failures rather than regenerated. The estimand and search-aware p-value
formula are unchanged by the execution backend.

The shared mechanics do not erase experimental design. Cross-sectional
analyses resample independent biological observations, destructive time
courses resample within condition-by-time cells, and repeated designs resample
complete subject trajectories. Duplicate trajectory draws receive new
replicate-subject identifiers so they remain statistically distinct copies of
one source trajectory. Unsupported exchangeability or insufficient distinct
assignments produces a typed abstention rather than a fallback sampling rule.

### Extending association methods

`AssociationStrategy` is the supported extension seam for additional metadata
interpretation methods. A registered adapter declares its supported sampling
designs and target types, estimand, cohort policy, diagnostics, abstention
statuses, scientific refit policy, and evidence version through
`association_contract()`. Simple adapters can inherit available-case cohort
preparation and index-based refitting; methods with different biological units
override `prepare_association()` and `refit_association()`.

The adapter does not construct atlas tables, digests, provenance, or public
result objects. `associate_metadata()` validates the contract, runs refits
through the shared future-backed execution layer, and normalizes the evidence
into the same atlas, proposal, and typed-abstention workflow used by the
built-in strategies. The practical author checklist and complete conformance
fixture are described in `docs/agents/association-strategy-authoring.md`.

## Produce an exploratory proposal

```{r proposal}
proposal <- propose_component(
  atlas,
  n_permutations = 49L,
  seed = 8001L
)
proposal_ranking(proposal)
proposal_provenance(proposal)
permutation <- proposal_permutation_evidence(proposal)
permutation
plot(proposal)
plot(permutation)
```

The proposal ranks the adjusted target effect because nuisance fields were
declared. Every nuisance-only residual permutation reconstructs a null score
and repeats the complete eligible-component search. The resulting p-value
therefore compares the observed maximum absolute effect with null maxima, not
with one component considered in isolation. It remains supporting exploratory
evidence and cannot promote a runner-up.

The default cross-sectional declaration treats canonical biological samples as
independent and exchangeable. If that assumption is not defensible, set
`exchangeability = "not_identifiable"` in `associate_metadata()`. Point
evidence remains available, while permutation returns
`permutation-not-identifiable` and no search-aware p-value.

If target intent or a defensible permutation structure is absent, the same call
returns a typed `ComponentAbstention` with
`permutation-not-identifiable`. A rank-deficient or target-confounded nuisance
design similarly preserves raw evidence while adjusted nomination abstains.

## Assess whether the proposed axis is recoverable

Association bootstraps quantify uncertainty conditional on a fitted Stage 1
basis. Axis-identifiability evidence asks a different question: whether the
complete discovery search recovers the same one-dimensional target structure
when the biological sampling units are resampled.

The assessment starts from the pre-decomposition `StateTransitionData`, the
frozen proposal, and the exact `PipelineConfig` used for discovery:

```{r identifiability-interface, eval=FALSE}
assessed <- assess_component_identifiability(
  data = raw_data,
  proposal = proposal,
  config = discovery_config,
  non_analytical_fields = "mouse_id",
  n_resamples = 499L,
  seed = 8301L
)

evidence <- proposal_identifiability(assessed)
evidence$recurrence_summary
evidence$target_recurrence
plot_component_identifiability(assessed)
plot_component_identifiability(assessed, view = "diagnostic")
# Complete nine-panel evidence record when a detailed audit is needed:
plot_component_identifiability(assessed, view = "audit")
```

Each replicate reruns the declared decomposition and complete eligible
component ranking. Cross-sectional observations are resampled as independent
biological units, destructive time courses within condition-by-time cells, and
longitudinal data as complete subject trajectories. Duplicated longitudinal
draws receive fresh subject identifiers.

The ordinary SVD strategy declares feature-loading cosine as its matching
geometry. A global one-to-one assignment retains the complete similarity
matrix, selected and competing assignments, margins, unmatched axes, sign
orientation, index recurrence, proposal-rank recurrence, subspace angles,
spectral gaps, source sampling units, and failed replicates. Sign correction
occurs only after matching; no Procrustes rotation is applied.

For an accepted single-component decomposition (`K = 1`), the same contract
remains explicit rather than inventing a comparison axis. The sole component
is matched to itself, its sign orientation is retained, and axis recurrence is
reported together with the one-dimensional subspace angle. Because there is no
competing assignment, local and global assignment margins are `NA`, not zero.
Requested, completed, abstaining, and failed resamples remain separately
counted in the evidence and in every plot view. For `K = 1`, the primary view
shows the distributions of loading agreement and one-dimensional subspace
rotation. The diagnostic view relates loading agreement to the independently
repeated biological-effect magnitude in each resample. The audit view retains
the complete nine-panel evidence record, including assignment-margin
non-applicability.
These quantities are
descriptive until the known-truth calibration in issue #67 freezes an
acceptance rule.

### AML-shaped longitudinal calibration control

The exported AML-shaped control joins the repeated-subject interpretation and
identifiability contracts in one disclosed workflow. It plants a dominant
collection-time direction and a weaker condition-by-time direction at the
irregular weeks supplied with the package. The control stores its feature
loadings, sample scores, and target orientation in
`K1AmlLongitudinalGroundTruth`. Versioned accessors expose that answer key and
the generator declaration without making internal `metadata()` fields part of
the public API. A user-supplied time grid is labelled as user supplied rather
than attributed to the packaged AML mapping.

```{r aml-k1-calibration, eval=FALSE}
aml_control <- k1_aml_longitudinal_calibration(
  subjects_per_condition = 12L,
  p = 100L,
  seed = 6700L,
  n_resamples = 49L
)

aml_longitudinal_control_truth(aml_control$control)
aml_longitudinal_control_info(aml_control$control)
aml_control$recovery
atlas_associations(aml_control$atlas)
proposal_identifiability(aml_control$identifiability)
plot_component_identifiability(aml_control$identifiability)
aml_control$stage2
```

Strategy selection comes from the returned `PipelineConfig`, so the calibration
uses the same registered plugin boundary as an ordinary analysis. The returned
loading cosines, orientation, principal angles, adjusted effects, and resampling
evidence are descriptive. No threshold or pass/fail decision is applied.

The independent-acceptance runner extends this calibration into a frozen
experiment without changing the scientific question. It evaluates 4, 7, and
12 mice per condition at 100, 1,000, and 10,000 expression features, with 100
independent replicates in each cell. Every successful replicate retains raw and
batch-adjusted associations, proposal or abstention evidence, and the typed
Stage 2 result, together with truth-derived recovery metrics and generator
provenance. Complete-mouse bootstrap evidence is retained when a component
proposal is reached; a valid proposal abstention instead records
identifiability as not assessed. Execution failures retain their reason and
remain in every requested-replicate denominator. The summary keeps
target-loading recovery, component-index recurrence, bootstrap completion, and
whole-cell pass rates separate rather than replacing them with one score.
`plot_k1_aml_acceptance_summary()` renders the synchronized cell and recovery
views; `scientific_caption()` returns their separate figure captions.

When a disclosed calibration run has finished, classify each requested
replicate before interpreting cell-level rates. This keeps failure to recover
the planted axis separate from a downstream model abstention and from missing
recovery evidence. The figure caption remains separate from the ggplot object,
and the published directory can be verified by replaying the assessment from
its stored source evidence.

```{r k1-calibration-outcomes, eval=FALSE}
assessment <- assess_k1_calibration_outcomes(results, tasks, protocol)
outcome_plot <- plot_k1_calibration_outcomes(assessment)
scientific_caption(outcome_plot)

artifact <- publish_k1_calibration_outcomes(
  artifact_root,
  results,
  tasks,
  protocol
)
verify_k1_calibration_outcomes(artifact)
```

Independent assessment is separated from calibration. Its seeds and criteria
are fixed before the outcomes are inspected and cannot be retuned from those
outcomes. All requested replicates contribute to the assessment, including
failed runs and valid abstentions. A preliminary resource-use pilot is
non-evidentiary and cannot contribute to the scientific result. Neither that
pilot, the calibration example, nor its visual proof supports a biological AML
claim.

`k1_revised_acceptance_targets()` builds the governed independent-assessment
graph without choosing cluster infrastructure. By default,
`controller = NULL` inherits the controller configured in the active
`targets` resources, such as the default supplied by `hprcc` in a Slurm
session. An explicit named controller remains available when a workflow needs
a reviewed resource override. Controller choice changes deployment only; it
does not change the frozen manifest, scientific commands, or RNG streams.

### Independent destructive-time-course operating evidence

Destructive sampling uses a different biological animal in every
condition-time cell. `k1_independent_time_course_template()` declares balanced
one-, two-, and three-animal cells, unequal cells, an isolated failed sequencing
library, and a missing internal cell. A failed library remains a recorded
removal from the intended design; it is never imputed or silently relabelled as
a smaller balanced experiment.

```{r k1-independent-time-calibration, eval=FALSE}
assessment <- run_k1_independent_time_course_calibration(
  replicates = 20L,
  p = 100L,
  seed = 18900L
)
operating_map <- plot_k1_independent_time_course_calibration(assessment)
operating_map
scientific_caption(operating_map)

artifact <- publish_k1_independent_time_course_calibration(
  artifact_root = "/absolute/shared/evidence",
  assessment = assessment
)
verify_k1_independent_time_course_calibration(artifact)
```

Panel A asks whether the planted feature-space axis was recovered. Panel B asks
whether the declared condition-by-time interaction was estimable after
recovery. A cross is an abstention, not a recovery probability of zero. The map
is evidence for the displayed synthetic regime only; it does not establish a
universal minimum sample size. For scheduler-backed work,
`k1_independent_time_course_calibration_targets()` places the complete
scientific call behind the existing targets and crew controller boundary, so a
local controller or an hprcc Slurm controller changes deployment rather than
the scientific graph.

### Incomplete repeated-subject operating evidence

Repeated-subject designs keep the mouse, condition, and observed time attached
to every retained observation. `k1_repeated_subject_template()` declares four
mice per condition observed at four intended times, then applies no loss, one
missed internal visit, terminal dropout in one disease mouse, or terminal
dropout in two disease mice. The removals remain part of the sampling audit;
they are not imputed or treated as independent animals.

```{r k1-repeated-subject-calibration, eval=FALSE}
assessment <- run_k1_repeated_subject_calibration(
  replicates = 20L,
  p = 100L,
  axis_resamples = 19L,
  seed = 19000L
)
operating_map <- plot_k1_repeated_subject_calibration(assessment)
operating_map
scientific_caption(operating_map)

artifact <- publish_k1_repeated_subject_calibration(
  artifact_root = "/absolute/shared/evidence",
  assessment = assessment
)
verify_k1_repeated_subject_calibration(artifact)
```

Panel A asks whether the planted feature-space target was recovered. Panel B
shows mean absolute loading cosine for that planted target under
complete-subject bootstrap refits; it is independent of whether metadata-based
component nomination succeeds.
Panel C asks whether the required random-intercept-plus-slope model remains
estimable after recovery. A cross is typed scientific non-estimability, not a
probability of zero. When the random slope is not identifiable, the method does
not fall back to a random-intercept-only model or observation-level
independence. `k1_repeated_subject_calibration_targets()` retains the same
scientific graph under local crew and hprcc Slurm controllers.

Before a known-truth calibration is frozen, the structured outcome is
`not-calibrated` and the evidence remains
`estimable-exploratory-only`. Singular-value gaps are visible diagnostics, not
effect weights or universal cutoffs. A later calibrated
`stable-subspace/no-stable-axis` or `no-stable-target-structure` outcome is a
successful scientific abstention and cannot be overridden to confirm an
individual component.

## Record the analyst decision

```{r confirmation, eval=FALSE}
confirmed <- confirm_component(
  assessed,
  index = assessed@recommended_component,
  decision = "accept",
  rationale = paste(
    "Accepted the unique effect-first component in this synthetic",
    "implementation example."
  )
)
confirmed
canonical_digest(confirmed)
```

`confirm_component()` is the enforced transition to a confirmed
`AnalysisSpecification`. Before calibration it records an explicitly
exploratory choice and forces `claim_intent = "exploratory"`; it does not label
the axis stable, validated, accepted as evidence, or scientifically supported.
After calibration, only digest-valid `stable-axis` evidence permits individual
component confirmation. Acceptance must use the recommended component. A
different effect-equivalent component still requires
`decision = "override"` and remains explicit. An override cannot bypass a
calibrated non-axis, `non-identifiable-design`,
`outside-calibrated-operating-region`, or `unique-winner-failure` outcome.

## Independent destructive-sampling time courses

A destructive time course observes different biological samples at each time
point. Time is known, but no subject trajectory is observed. Declare that
design explicitly with `independent_time_course()`. Do not invent subject IDs
or use `longitudinal()` for these data.

The synthetic example below plants a condition-by-time divergence across two
components. It uses three independent samples in every condition-by-time cell.

```{r destructive-time-course}
time <- rep(c(0, 2, 4), each = 6L)
condition <- factor(
  rep(rep(c("control", "treatment"), each = 3L), 3L),
  levels = c("control", "treatment")
)
batch <- factor(rep(c("run_1", "run_2", "run_1"), 6L))
primary_time <- sprintf("time_sample_%02d", seq_along(time))
assay_time <- sprintf("rna_time_%02d", seq_along(time))

time_data <- StateTransitionData(
  experiments = list(
    rna = SummarizedExperiment::SummarizedExperiment(
      assays = list(logcounts = matrix(
        seq_len(4L * length(time)),
        nrow = 4L,
        dimnames = list(sprintf("gene_%02d", 1:4), assay_time)
      ))
    )
  ),
  colData = S4Vectors::DataFrame(
    condition = condition,
    day = time,
    batch = batch,
    row.names = primary_time
  ),
  sampleMap = S4Vectors::DataFrame(
    assay = factor(rep("rna", length(time)), levels = "rna"),
    primary = primary_time,
    colname = assay_time
  )
)
time_data <- declare_sampling_design(
  time_data,
  independent_time_course(time = "day", time_unit = "days")
)
time_metadata <- S4Vectors::metadata(time_data)
time_metadata$stage1 <- DecompositionResult(
  V_star = c(1, 0, 0, 0),
  sigma = 1,
  coords = list(seq_along(time)),
  V_k = diag(4)[, 1:2, drop = FALSE],
  sigma_k = matrix(c(2, 1), nrow = 1L),
  coords_k = list(cbind(
    PC1 = as.numeric(scale(time)),
    PC2 = as.numeric(scale(
      time * as.integer(condition == "treatment") +
        rep(c(-0.15, 0, 0.15), 6L)
    ))
  )),
  k = 2L
)
S4Vectors::metadata(time_data) <- time_metadata

time_specification <- analysis_specification(
  id = "synthetic-time-divergence",
  target_field = "condition",
  target_type = "binary",
  reference_level = "control",
  comparison_level = "treatment",
  nuisance_fields = "batch"
)
time_atlas <- associate_metadata(
  time_data,
  specification = time_specification,
  dataset_id = "synthetic-destructive-time-course",
  n_resamples = 49L,
  seed = 8101L
)
atlas_associations(time_atlas)
plot(time_atlas)
```

The registered ordinary linear model is
`score_std ~ condition * time_scaled + nuisance_terms`. Scores have a
deterministic orientation and SD scale. Observed study time is transformed
deterministically to the study's 0--1 interval, and the interaction coefficient
is the proposal-eligible effect. Raw and adjusted variants remain separate.
The stored evidence records the analysis cohort, formula, model-matrix rank,
time transformation, cell counts, bootstrap failures, engine controls, and
provenance.

Both conditions must occur at two or more overlapping times, every relevant
condition-by-time cell must contain independent replication, the design matrix
must be full rank, and the number of observations must exceed its degrees of
freedom. A failed gate produces a typed abstention. Bootstrap resampling occurs
within condition-by-observed-time cells and retains the observed grid and cell
counts. Fixed observed times are never permuted. Unadjusted proposals permute
condition labels within time; adjusted proposals use nuisance-only residual
permutation within time.

```{r destructive-time-course-proposal}
time_proposal <- propose_component(
  time_atlas,
  n_permutations = 49L,
  seed = 8201L
)
proposal_ranking(time_proposal)
proposal_permutation_evidence(time_proposal)
plot(time_proposal)
```

This declaration is distinct from two other designs:

- repeated subjects require complete subject trajectories to be resampled and
  a subject-aware random-effects association strategy;
- ordered cross-sectional states express a declared ordering, not elapsed
  observed time or within-subject change.

## Repeated-subject time courses

Repeated measurements declare both the subject identifier and observed study
time. The registered strategy estimates the standardized condition-by-time
interaction with correlated subject-specific random intercepts and time slopes:
`score_std ~ condition * time_scaled + nuisance_terms +
(1 + time_scaled | subject)`.

```{r repeated-subject-time-course}
subject_id <- rep(sprintf("mouse_%02d", 1:16), each = 4L)
subject_index <- rep(1:16, each = 4L)
repeated_time <- rep(0:3, times = 16L)
repeated_condition <- factor(
  rep(rep(c("control", "treatment"), each = 8L), each = 4L),
  levels = c("control", "treatment")
)
subject_offset <- 0.12 * sin(subject_index)
subject_slope <- 0.06 * ((subject_index - 1L) %% 8L - 3.5)
repeated_pc1 <- subject_offset +
  (0.5 + subject_slope) * repeated_time +
  1.4 * (repeated_condition == "treatment") * repeated_time +
  0.05 * sin(seq_along(repeated_time))
repeated_pc2 <- 0.1 * cos(subject_index) +
  (0.8 - subject_slope) * repeated_time +
  0.05 * cos(seq_along(repeated_time))
repeated_primary <- sprintf("repeated_sample_%03d", seq_along(repeated_time))
repeated_assay <- sprintf("repeated_rna_%03d", seq_along(repeated_time))

repeated_data <- StateTransitionData(
  experiments = list(
    rna = SummarizedExperiment::SummarizedExperiment(
      assays = list(logcounts = matrix(
        seq_len(5L * length(repeated_time)),
        nrow = 5L,
        dimnames = list(sprintf("gene_%02d", 1:5), repeated_assay)
      ))
    )
  ),
  colData = S4Vectors::DataFrame(
    condition = repeated_condition,
    day = repeated_time,
    mouse_id = subject_id,
    batch = factor(ifelse(subject_index %% 2L, "run_1", "run_2")),
    row.names = repeated_primary
  ),
  sampleMap = S4Vectors::DataFrame(
    assay = factor(
      rep("rna", length(repeated_time)),
      levels = "rna"
    ),
    primary = repeated_primary,
    colname = repeated_assay
  )
)
repeated_data <- declare_sampling_design(
  repeated_data,
  longitudinal(
    subject_id = "mouse_id",
    time = "day",
    time_unit = "days"
  )
)
repeated_metadata <- S4Vectors::metadata(repeated_data)
repeated_metadata$stage1 <- DecompositionResult(
  V_star = c(1, 0, 0, 0, 0),
  sigma = 1,
  coords = list(repeated_pc1),
  V_k = diag(5)[, 1:2, drop = FALSE],
  sigma_k = matrix(c(2, 1), nrow = 1L),
  coords_k = list(cbind(
    PC1 = repeated_pc1,
    PC2 = repeated_pc2
  )),
  k = 2L
)
S4Vectors::metadata(repeated_data) <- repeated_metadata

repeated_specification <- analysis_specification(
  id = "synthetic-repeated-divergence",
  target_field = "condition",
  target_type = "binary",
  reference_level = "control",
  comparison_level = "treatment",
  nuisance_fields = "batch"
)
repeated_atlas <- associate_metadata(
  repeated_data,
  specification = repeated_specification,
  non_analytical_fields = "mouse_id",
  dataset_id = "synthetic-repeated-time-course",
  n_resamples = 49L,
  seed = 8202L
)
repeated_proposal <- propose_component(
  repeated_atlas,
  n_permutations = 49L,
  seed = 8203L
)
plot(repeated_atlas)
plot(repeated_proposal)
```

Repeated-subject atlas and proposal plots label component panels A, B, and so
on in reading order. Their separate scientific captions use the same letters
to explain each component-specific trajectory and interaction estimate.

The complete subject trajectory is the inferential and resampling unit.
Bootstrap draws occur within condition and receive fresh subject identifiers,
so a duplicated draw remains one trajectory rather than pseudoreplicated
observations. Permutation reassigns condition only at subject level and keeps
observed times fixed. Missing required observations exclude the complete
subject trajectory from a model; the recorded study-time range is not silently
contracted.

At least three usable observed times per subject, independent between-subject
replication in each condition, and a full-rank fixed-effects design are
required. Singular random-effects covariance and optimizer non-convergence are
distinct typed abstentions with native diagnostics. Neither outcome triggers a
random-intercept-only or independent-observation fallback. Individual
trajectories, condition-level fitted divergence, sampling support, dropout,
uncertainty, fit failures, and abstention remain visible in the canonical
plots.

All three supported sampling designs cross the same package-owned evidence
boundary before the atlas is constructed. `atlas_evidence_contract()` returns
the same inspection fields for each design: a module version and sampling
design, normalized row counts, deterministic table digests, cohort summaries,
and explicit cohort membership. The common shape makes downstream evidence
auditable without treating condition-by-time cells and subject trajectories as
exchangeable:

```{r evidence-contracts}
lapply(
  list(
    cross_sectional = atlas,
    independent_time_course = time_atlas,
    repeated_subject = repeated_atlas
  ),
  function(x) {
    contract <- atlas_evidence_contract(x)
    list(
      version = contract$version,
      sampling_design = contract$sampling_design,
      row_counts = contract$row_counts
    )
  }
)
```

## Scientific captions are separate publication artifacts

User-facing landscapeR renderers return ordinary `ggplot` objects. Migrated
renderers attach a deterministic caption derived from the same typed evidence
used by the graphic; retrieve it with `scientific_caption(plot)`. The package
does not draw this text inside the plot, so Quarto, R Markdown, and manuscript
systems can render it as a true figure caption. Caption templates preserve
caller-declared experiment, layer, target, level, time, subject, and nuisance
labels exactly. They format stored estimands, uncertainty, thresholds,
missingness, and claim boundaries without recalculating scientific results.

The same boundary is available directly through `visual_evidence()`. It returns
a validated `VisualEvidenceView` used by the canonical ggplot renderer:

```{r visual-evidence-view, eval=FALSE}
view <- visual_evidence(proposal)

visual_evidence_surface(view)
visual_evidence_state(view)
visual_evidence_observations(view)
visual_evidence_summaries(view)
visual_evidence_diagnostics(view)
visual_evidence_caption(view)
```

Design-specific display tables are retrieved by their recorded names with
`visual_evidence_display()`. They contain already-computed trajectories,
interval labels, missing-cell markers, dropout endpoints, or fitted diagnostic
curves as applicable. Plotting code does not inspect private provenance or
reconstruct those scientific results. The view can support a future compatible
interactive adapter, but ggplot2 remains the canonical reproducible output and
interactive software cannot alter a proposal or abstention.

Stage 1 and Stage 2 use the same consumer contract. Their digest-bound
`StagePlotEvidence` remains an internal storage payload for compatibility, and
`visual_evidence()` adapts it to `VisualEvidenceView` before rendering. Missing,
invalid or stale display evidence produces a captioned unavailable plot rather
than changing the success status of decomposition or dynamics estimation.

## Registering additional methods

Method authors register an R closure under one `contract:name` key with
`register_strategy()`. Keys are write-once by default: a different constructor
under an existing key raises a typed collision error and leaves the accepted
method unchanged, while re-registering a fingerprint-equivalent constructor
with the same code and behavior-relevant captured state is an idempotent no-op.
Primitive functions are unsupported because they cannot meet the inspectable
fingerprint contract.

A deliberate correction uses `replace = TRUE` with a non-empty rationale. The
accepted registration and any replacement remain inspectable with
`strategy_registration_history()`, including chained constructor fingerprints
and the replacement rationale. A registered closure whose behavior-relevant
captured state has changed must be stabilized and registered from a fresh
session rather than silently replacing its recorded identity.

## Current boundary

These paths do not reinterpret destructive time courses as longitudinal
subjects or claim axis stability before calibration. Association-bootstrap
intervals quantify uncertainty while holding the fitted Stage 1 basis fixed;
axis-identifiability evidence separately refits Stage 1 and the full component
search. The current repeated-subject strategy is linear and frequentist;
nonlinear, marginal, robust, or Bayesian trajectory models require separately
registered strategies. A real-data confirmation is a documented analyst
decision, not proof that the selected coordinate is biologically true.
