---
title: "Cross-sectional binary component interpretation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Cross-sectional binary component interpretation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 4,
  fig.height = 4,
  dpi = 120
)
library(landscapeR)
```

This workflow is the first complete interpretation path in `landscapeR`. It
supports one molecular layer, independent cross-sectional biological samples,
and a binary biological target. Every result remains exploratory.

The workflow deliberately separates four actions:

1. `associate_metadata()` calculates and stores eligible associations.
2. `plot()` renders the stored evidence without changing it.
3. `propose_component()` ranks only by absolute signed rank-biserial effect.
4. `confirm_component()` records a separate human decision and rationale.

## Synthetic worked example

The example uses planted synthetic truth to demonstrate software behavior. It
is not biological evidence and does not calibrate a universal decision rule.

```{r data}
primary <- sprintf("sample_%02d", 1:8)
assay_ids <- sprintf("rna_%02d", 1:8)
std <- StateTransitionData(
  experiments = list(
    rna = SummarizedExperiment::SummarizedExperiment(
      assays = list(logcounts = matrix(
        seq_len(32L),
        nrow = 4L,
        dimnames = list(sprintf("gene_%02d", 1:4), assay_ids)
      ))
    )
  ),
  colData = S4Vectors::DataFrame(
    condition = factor(
      rep(c("control", "treatment"), each = 4L),
      levels = c("control", "treatment")
    ),
    mouse_id = sprintf("mouse_%02d", 1:8),
    row.names = primary
  ),
  sampleMap = S4Vectors::DataFrame(
    assay = factor(rep("rna", 8L), levels = "rna"),
    primary = primary,
    colname = assay_ids
  )
)
std <- declare_sampling_design(std, cross_sectional())
std_metadata <- S4Vectors::metadata(std)
std_metadata$stage1 <- DecompositionResult(
  V_star = c(1, 0, 0, 0),
  sigma = 1,
  coords = list(1:8),
  V_k = diag(4)[, 1:2, drop = FALSE],
  sigma_k = matrix(c(2, 1), nrow = 1L),
  coords_k = list(cbind(
    PC1 = 1:8,
    PC2 = rep(c(-1, 1), 4L)
  )),
  k = 2L
)
S4Vectors::metadata(std) <- std_metadata
```

## Build and inspect the atlas

```{r atlas}
atlas <- associate_metadata(
  std,
  non_analytical_fields = "mouse_id",
  dataset_id = "synthetic-binary-control"
)
atlas_associations(atlas)
head(atlas_observations(atlas))
atlas_exclusions(atlas)
atlas_provenance(atlas)
plot(atlas)
```

Binary factor levels define reference then comparison. Character and logical
targets use deterministic lexical order. Identifier-like fields and declared
non-analytical fields are excluded with recorded reasons. Missing values use
available cases, and the tidy association table records available counts,
missing counts, and tied score counts.

The atlas is a versioned `MetadataAssociationAtlas`. Its digest binds the
canonical metadata and sample mapping, raw typed distributions, frozen Stage 1
state space, computation tier, strategy, package version, and sampling design.

## Produce an exploratory proposal

```{r proposal}
proposal <- propose_component(atlas, target = "condition")
proposal_ranking(proposal)
proposal_provenance(proposal)
plot(proposal)
```

The proposal consumes stored atlas rows. It does not use p-values, singular
values, stability, or a plot to rerank components. An exact top-effect tie
returns a typed `ComponentAbstention` instead of choosing a component by index.

## Record human confirmation

```{r confirmation}
confirmed <- confirm_component(
  proposal,
  index = proposal@recommended_component,
  decision = "accept",
  rationale = paste(
    "Accepted the unique effect-first component in this synthetic",
    "implementation example."
  )
)
confirmed
canonical_digest(confirmed)
```

`confirm_component()` is the enforced transition to a confirmed
`AnalysisSpecification`. Acceptance must use the recommended component. A
different ranked component requires `decision = "override"` and remains
explicit in the specification. An abstention cannot be confirmed.

## Current boundary

This implementation does not yet cover adjusted associations, continuous or
ordered targets, longitudinal sampling, repeated subjects, multiple molecular
layers, or stability-calibrated evidence. Those require separate registered
strategies and validation. A real-data confirmation is a recorded human
judgement, not proof that the selected coordinate is biologically true.
