---
title: "Cross-sectional component interpretation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Cross-sectional component interpretation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 4,
  fig.height = 4,
  dpi = 120
)
library(landscapeR)
```

This workflow interprets one molecular layer measured on independent
cross-sectional biological samples. Binary, continuous, and ordered targets
have registered association strategies. Unordered multilevel fields remain
visible as descriptive omnibus evidence but cannot nominate a component.
Every result remains exploratory.

The workflow separates four actions:

1. `associate_metadata()` stores raw, adjusted, and resampled evidence.
2. `plot()` renders the stored decision surface without changing it.
3. `propose_component()` ranks only the predeclared effect and may compute a
   search-aware permutation null.
4. `confirm_component()` records a separate human decision and rationale.

## Synthetic worked example

The example demonstrates software behavior with synthetic coordinates. It is
not biological evidence and does not calibrate a universal decision rule.

```{r data}
primary <- sprintf("sample_%02d", 1:8)
assay_ids <- sprintf("rna_%02d", 1:8)
std <- StateTransitionData(
  experiments = list(
    rna = SummarizedExperiment::SummarizedExperiment(
      assays = list(logcounts = matrix(
        seq_len(32L),
        nrow = 4L,
        dimnames = list(sprintf("gene_%02d", 1:4), assay_ids)
      ))
    )
  ),
  colData = S4Vectors::DataFrame(
    condition = factor(
      rep(c("control", "treatment"), each = 4L),
      levels = c("control", "treatment")
    ),
    severity = c(1, 2, 2, 4, 5, 6, 7, 8),
    state = ordered(
      c("early", "early", "middle", "middle",
        "late", "late", "late", "end"),
      levels = c("early", "middle", "late", "end")
    ),
    subtype = factor(rep(c("A", "B", "C", "D"), each = 2L)),
    batch = rep(c("run_1", "run_2"), 4L),
    mouse_id = sprintf("mouse_%02d", 1:8),
    row.names = primary
  ),
  sampleMap = S4Vectors::DataFrame(
    assay = factor(rep("rna", 8L), levels = "rna"),
    primary = primary,
    colname = assay_ids
  )
)
std <- declare_sampling_design(std, cross_sectional())
std_metadata <- S4Vectors::metadata(std)
std_metadata$stage1 <- DecompositionResult(
  V_star = c(1, 0, 0, 0),
  sigma = 1,
  coords = list(1:8),
  V_k = diag(4)[, 1:2, drop = FALSE],
  sigma_k = matrix(c(2, 1), nrow = 1L),
  coords_k = list(cbind(
    PC1 = c(4, 3, 2, 1, 1, 2, 3, 4),
    PC2 = 1:8
  )),
  k = 2L
)
S4Vectors::metadata(std) <- std_metadata
```

## Declare intent, then build the atlas

The specification owns the target, direction, nuisance fields, and run
identity. Numerical spacing is explicit here because `severity` is declared as
continuous. Ordered factor labels instead use their declared level order and
never imply unequal spacing.

```{r atlas}
specification <- analysis_specification(
  id = "synthetic-severity",
  target_field = "severity",
  target_type = "continuous",
  continuous_direction = "increasing",
  nuisance_fields = "batch"
)

atlas <- associate_metadata(
  std,
  specification = specification,
  non_analytical_fields = "mouse_id",
  dataset_id = "synthetic-cross-sectional-control",
  n_resamples = 49L,
  seed = 7001L
)
atlas_associations(atlas)
atlas_exclusions(atlas)
plot(atlas)
```

The table keeps unadjusted and adjusted rows separate. Each row records its
complete-cohort and design digests, missingness, tied score and target counts,
bootstrap interval, failed-resample count, raw p-value, and within-field BH
q-value. Bootstrap draws preserve discrete target and nuisance-cell counts and
resample complete independent biological units rather than features or
technical replicates.

Raw points, a black monotone-constrained fit, and a red flexible smoother
expose the predeclared monotone relationship. A reversal in level-wise medians adds a
`possible-nonmonotone-association` warning. The warning and smoother are
descriptive and cannot rerank components.

## Produce an exploratory proposal

```{r proposal}
proposal <- propose_component(
  atlas,
  n_permutations = 49L,
  seed = 8001L
)
proposal_ranking(proposal)
proposal_provenance(proposal)
permutation <- proposal_permutation_evidence(proposal)
permutation
plot(proposal)
plot(permutation)
```

The proposal ranks the adjusted target effect because nuisance fields were
declared. Every nuisance-only residual permutation reconstructs a null score
and repeats the complete eligible-component search. The resulting p-value
therefore compares the observed maximum absolute effect with null maxima, not
with one component considered in isolation. It remains supporting exploratory
evidence and cannot promote a runner-up.

The default cross-sectional declaration treats canonical biological samples as
independent and exchangeable. If that assumption is not defensible, set
`exchangeability = "not_identifiable"` in `associate_metadata()`. Point
evidence remains available, while permutation returns
`permutation-not-identifiable` and no search-aware p-value.

If target intent or a defensible permutation structure is absent, the same call
returns a typed `ComponentAbstention` with
`permutation-not-identifiable`. A rank-deficient or target-confounded nuisance
design similarly preserves raw evidence while adjusted nomination abstains.

## Record human confirmation

```{r confirmation}
confirmed <- confirm_component(
  proposal,
  index = proposal@recommended_component,
  decision = "accept",
  rationale = paste(
    "Accepted the unique effect-first component in this synthetic",
    "implementation example."
  )
)
confirmed
canonical_digest(confirmed)
```

`confirm_component()` is the enforced transition to a confirmed
`AnalysisSpecification`. Acceptance must use the recommended component. A
different ranked component requires `decision = "override"` and remains
explicit. An abstention cannot be confirmed.

## Current boundary

This path does not reinterpret destructive time courses as longitudinal
subjects, support repeated-subject models, combine molecular layers, or claim
axis stability. Those require separate registered strategies and later
calibration. Bootstrap intervals here quantify association uncertainty while
holding the fitted Stage 1 basis fixed. A real-data confirmation is a recorded
human judgement, not proof that the selected coordinate is biologically true.
