---
title: "Development log"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Development log}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 8,
  fig.height = 5,
  dpi = 120
)
library(landscapeR)
```

> **This is a living implementation record, not the package schedule or an
> acceptance-evidence report.** The root
> [ROADMAP](https://github.com/drejom/landscapeR/blob/main/ROADMAP.md) is the
> single scheduling authority. This article shows current user-visible behavior
> and links to implementation/evidence records. Acceptance evidence is published
> only from immutable, hash-verified benchmark artifacts.

---

## Current implementation status

| Stage | Name | Status |
|---|---|---|
| Architecture | S4 contracts, registry, container, provenance | ✅ Implemented and CI-tested |
| 0 | Synthetic controls (K=1/K≥2 subspace, exact-stationary double well) | ⚠️ Calibration controls implemented; frozen K=1 recovery/negative-control ladder remains #51 |
| 1 | Single-omic-layer registered SVD | ⚠️ Implemented and visually demonstrated; scientific acceptance remains blocked by #51 and #67 |
| 1 | Comparative decomposition — legacy equal-feature strategies | ⚠️ Implemented; not evidence for heterogeneous multi-omic fitting |
| 1 | Heterogeneous C1/C2 prototype and frozen comparison protocol | ⚠️ Full synthetic run complete; C2 won calibration but failed frozen holdout, so no baseline is accepted and #24 remains blocked |
| 1+ | Component gallery and legacy `project_into()` workflow | ⚠️ Exploratory/legacy; #24 is blocked on the selected heterogeneous strategy |
| 2 | KDE log-density inversion and constrained polynomial smooth | ⚠️ Implemented; ADR 0002 remains provisional-accepted |

Future sequencing, dependencies, and parked capabilities are maintained only in
[ROADMAP.md](https://github.com/drejom/landscapeR/blob/main/ROADMAP.md), so this
article does not duplicate a second planned-work list.

### Biological instantiations

| Dataset | Modality | Status |
|---|---|---|
| AML/CML (Cbfb-MYH11, BCR-ABL) | mRNA, miRNA | Historical exploratory context only; raw data are not in this repository and no claim is current |
| Pogona TSD | mRNA | Planned only; requires a validated geometry/applicability assessment |
| Islet/diabetes multi-omic | mRNA + ATAC + genotype | Planned only; rank-deficient genotype layer is a required Stage 1 stress case |

---

## Analysis intent lifecycle (AnalysisSpecification v2)

A component decision now **resolves rather than replaces** its target. Every
specification retains the target field and its neutral contrast/order/direction;
confirmation adds the frozen-basis component and human decision provenance.

```{r analysis-specification-v2}
proposal_id <- strrep("a", 64L)

spec_draft <- analysis_specification(
  id = "condition-axis",
  target_field = "condition",
  target_type = "binary",
  reference_level = "control",
  comparison_level = "disease"
)

spec_proposal <- analysis_specification(
  id = "condition-axis",
  target_field = "condition",
  target_type = "binary",
  reference_level = "control",
  comparison_level = "disease",
  lifecycle = "proposal",
  proposal_digest = proposal_id
)

spec_confirmed <- analysis_specification(
  id = "condition-axis",
  target_field = "condition",
  target_type = "binary",
  reference_level = "control",
  comparison_level = "disease",
  lifecycle = "confirmed",
  selected_component = 2L,
  proposal_digest = proposal_id,
  proposal_decision = "accepted",
  analyst_rationale = "Accepted the predeclared top-ranked target axis."
)

data.frame(
  lifecycle = c(spec_draft@lifecycle,
                spec_proposal@lifecycle,
                spec_confirmed@lifecycle),
  target = rep("control -> disease", 3L),
  proposal = c("absent", "present", "retained"),
  selected_component = c(NA_integer_, NA_integer_,
                         spec_confirmed@selected_component),
  decision = c("absent", "absent", spec_confirmed@proposal_decision)
)
```

Legacy v1 objects migrate only through `migrate_analysis_specification()`.
Target-only v1 objects require explicit direction; component-only v1 objects
also require the missing target and proposal decision record. Migration fails
rather than inventing intent or dropping the legacy component. See the
[complete v2 and migration contract](https://github.com/drejom/landscapeR/blob/main/docs/specs/analysis-specification.md).

---

## Stage 0 — Synthetic validation

The pipeline is tested against two synthetic ground truths.

### Two-group control

`synthetic_control()` generates two groups of samples with a planted signal
on the first component. Used to verify Stage 1 recovers the correct axis.

```{r synthetic-control}
library(landscapeR)

std <- synthetic_control(n = 40L, p = 500L, K = 2L, signal = 30, seed = 1L)
ctor <- get_strategy("Decomposer", "hogsvd_averaged")
std1 <- suppressWarnings(decompose(ctor(), std))@value

plot_spectrum(std1)
```

> The metadata-coloured component gallery is temporarily omitted here because
> its canonical MAE-level metadata alignment is being repaired in
> [#54](https://github.com/drejom/landscapeR/issues/54). The spectrum above is
> descriptive only and does not select a target biological axis.

### K=1 single-omic-layer SVD and double-well calibration (#50)

Before issue #50, the only runnable development-log path duplicated a
single-omic-layer coordinate assay so it could enter a K≥2 HO-GSVD strategy.
That workaround obscured the actual capability boundary and recorded the wrong
strategy for a K=1 analysis.

| Condition | Before #50 | After #50 |
|---|---|---|
| K=1 synthetic subspace | Rejected by the generator | First-class control with planted subspace truth |
| Stage 1 strategy | Duplicate the assay and call HO-GSVD | Registered exactly-one-omic-layer `svd` |
| Double-well observations | Thinned trajectory in the older estimator-level control | Independent exact-stationary draws for the K=1 end-to-end calibration control |
| Evidence status | Ambiguous development workaround | Explicitly `non_evidentiary_calibration` |

The current development path now keeps K=1 explicit from generation through
Stage 1 and the cross-sectional Stage 2 estimator:

```{r k1-double-well, fig.cap=c("Descriptive evidence: empirical density of the planted cross-sectional coordinate supplied to the single-omic-layer expression control.", "Hypothesis-conditioned interpretation: the fitted calibration-only quasi-potential; critical-point classification is deliberately omitted because uncertainty is not yet estimated.")}
std_k1 <- synthetic_k1_double_well_control(
  n = 200L, p = 100L, noise_sd = 0.03, seed = 50L
)

svd_ctor <- get_strategy("Decomposer", "svd")
std_k1_s1 <- suppressWarnings(decompose(svd_ctor(), std_k1))@value

coordinate_k1 <- colData(std_k1)$x_coord

empirical_density <- ggplot2::ggplot(
  data.frame(coordinate = coordinate_k1),
  ggplot2::aes(x = coordinate)
) +
  ggplot2::geom_density(fill = "#67A9CF", alpha = 0.45, linewidth = 0.8) +
  ggplot2::geom_rug(alpha = 0.35) +
  ggplot2::labs(
    title = "Descriptive evidence: planted cross-sectional coordinate density",
    x = "State-transition coordinate",
    y = "Density"
  ) +
  ggplot2::theme_bw(base_size = 11)
print(empirical_density)

dynamics_ctor <- get_strategy("DynamicsEstimator", "kde_logdensity")
std_k1_s2 <- estimate_dynamics(dynamics_ctor(), std_k1_s1)@value
plot_potential(std_k1_s2, show_critical_points = FALSE) +
  ggplot2::labs(
    subtitle = paste(
      "Critical-point classification omitted — uncertainty is not yet",
      "estimated; calibration-only output"
    )
  )
```

**Cold-reader conclusion:** the single-omic-layer expression control now passes
through the registered SVD strategy and produces the expected two-well
quasi-potential without fabricating a second omic layer. The public synthetic
input distribution remains visible before the fitted interpretation, and the
plot omits critical-point classification because uncertainty is not yet estimated.

This is implementation and calibration proof, not acceptance evidence. The
frozen K=1 thinness, negative-control, and independent acceptance protocol is
tracked in [#51](https://github.com/drejom/landscapeR/issues/51); the
AML-grounded target/confounder control is [#67](https://github.com/drejom/landscapeR/issues/67).

```{r k1-calibration-table}
k1_calibration <- suppressWarnings(k1_double_well_calibration(
  n = 200L, p = 100L, noise_sd = 0.03, seed = 50L
))

knitr::kable(data.frame(
  diagnostic = c(
    "Evidence status",
    "Stage 1 strategy",
    "Subspace angle (degrees)",
    "Mean well-location error",
    "Barrier-location error",
    "Barrier-height error"
  ),
  value = c(
    k1_calibration$evidence_status,
    k1_calibration$decomposer,
    format(round(k1_calibration$subspace_angle_deg, 3), nsmall = 3),
    format(round(k1_calibration$well_error, 3), nsmall = 3),
    format(round(k1_calibration$barrier_error, 3), nsmall = 3),
    format(round(k1_calibration$barrier_height_error, 3), nsmall = 3)
  )
), col.names = c("Diagnostic", "Observed calibration output"))
```

The table deliberately contains no pass, acceptance, or eligibility judgement.
Those decisions belong to the frozen independent protocol, not this disclosed
calibration run.

---

## Historical AML/CML exploratory context

Earlier exploratory work used AML/CML cohorts as a translation target for the
Rockne--Frankhouser literature. The raw matrices and generated local
preprocessing objects are deliberately **not** included in this public
repository, so the old AML chunks are removed rather than presented as runnable
analysis. A minimal mapping of already-public mouse sample identifiers to the
checksum-pinned authoritative `sample_weeks` values is included solely to make
data lineage testable.

The corrected GSE133642 preparation keeps the source-study roles distinct from
landscapeR's use of both matrices as real biological validation data:

| Prepared layer | Source-study role | Observations | landscapeR role |
|---|---|---:|---|
| `mrna_primary_2018` | 2018 training cohort | 132 | future exploratory discovery basis |
| `mrna_supp_2016` | 2016 validation cohort 1 | 101 | later hostile batch/time-confounded projection test |

Its longitudinal declaration names only `mouse_id` and exact numeric
`sample_weeks` (unit: weeks). Categorical labels are inert, and no endpoint or
event semantics are inferred.

These data are not evidence for the package's present claims. Under the
[evidence-first validation plan](https://github.com/drejom/landscapeR/blob/main/docs/plans/implementation-plan-stage0-validation.md),
real-data Stage 2 inference is blocked until the frozen generic Stage 0 ladder
has passed, ADR 0002 has finalized thresholds, and a separate applicability and
confirmation workflow has been specified. Any future AML/CML article will
include a public data manifest, a fixed analysis specification, and the required
confirmation/claim-status diagnostics.

---

## Evidence status

The accepted Stage 1 v2 protocol compared two prototype estimators on
synthetic, heterogeneous-feature controls. Its completed, hash-verified full
artifact selected `C2_block_scaled_svd` in calibration, but C2 failed the
frozen holdout acceptance thresholds. Consequently no Stage 1 baseline is
accepted and #24 remains blocked. The [Stage 1 synthetic evidence article](stage1-evidence.html)
renders the immutable artifact; neither C1 substitution nor C2 tuning is
permitted from the consumed v2 holdout.

Similarly, the Stage 2 implementation is an auditable estimator, not yet a
validated biological inference engine. See ADR 0002 and the Stage 0 protocol
for its current claim boundary.

---

## Architecture notes

The package uses a **contract/registry pattern** — all stage algorithms are
S4 virtual classes with generics (`decompose`, `estimate_dynamics`). Concrete
strategies are registered by name and retrieved at runtime:

```{r registry}
list_strategies("Decomposer")
list_strategies("DynamicsEstimator")
```

This means adding a new algorithm (e.g. Kempf HO-GSVD, neural SDE) is a
`register_strategy()` call with no changes to calling code or tests.

---

## Implementation index

- **2026-07-14 — K=1 foundation and visual landing-proof policy:**
  [#50](https://github.com/drejom/landscapeR/issues/50) adds the registered
  single-omic-layer SVD and calibration control;
  [#68](https://github.com/drejom/landscapeR/issues/68) makes the pull request
  the canonical visual transition record. The full before/after proof is
  co-located with the diff in [PR #69](https://github.com/drejom/landscapeR/pull/69).
  This remains implementation work, not K=1 acceptance evidence.

*Last updated: 2026-07-14. Documentation site: <https://drejom.github.io/landscapeR/>.*
