---
title: "Development log"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Development log}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 8,
  fig.height = 5,
  dpi = 120
)
library(landscapeR)
```

> **This is a living document, not an acceptance-evidence report.** It tracks
> implementation progress and links to the governing protocols. Acceptance
> evidence is published only from immutable, hash-verified benchmark artifacts.

---

## Roadmap

### Implemented

| Stage | Name | Status |
|---|---|---|
| Architecture | S4 contracts, registry, container, provenance | ✅ Implemented and CI-tested |
| 0 | Synthetic controls (two-group SVD, double-well Langevin) | ⚠️ Implemented; full recovery/negative-control ladder not yet executed |
| 1 | Comparative decomposition — legacy equal-feature strategies | ⚠️ Implemented; not evidence for heterogeneous multi-omic fitting |
| 1 | Heterogeneous C1/C2 prototype and frozen comparison protocol | ⚠️ Full synthetic run complete; C2 won calibration but failed frozen holdout, so no baseline is accepted and #24 remains blocked |
| 1+ | Component gallery and legacy `project_into()` workflow | ⚠️ Exploratory/legacy; #24 is blocked on the selected heterogeneous strategy |
| 2 | KDE log-density inversion and constrained polynomial smooth | ⚠️ Implemented; ADR 0002 remains provisional-accepted |

### Planned

| Stage | Name | Notes |
|---|---|---|
| 0.5 | Archetype classifier | `ranger`-based; assigns samples to known archetypes |
| 0.75 | Distributional fit assessment | SBI backend via `reticulate`; replaces ad-hoc BC threshold |
| 1 | New heterogeneous-candidate decision | Requires a new ADR and frozen protocol; v2 holdout cannot be reused for tuning |
| 1 | HO-GSVD rank-deficiency handling | Kempf algorithm; candidate only after the current comparison decision |
| 2 | Full Stage 0 recovery and negative-control ladder | Establish ADR 0002 thresholds and pass rates |
| 2 | Minimum barrier height filter | Candidate only after Stage 0 evidence |
| 2 | Fokker-Planck time-to-transition | Requires a separately validated longitudinal strategy |
| 2 | Survival prediction | Requires validated real-data applicability and confirmation |
| — | Real-data confirmation workflows | Only after the generic Stage 0 and applicability gates pass |

### Biological instantiations

| Dataset | Modality | Status |
|---|---|---|
| AML/CML (Cbfb-MYH11, BCR-ABL) | mRNA, miRNA | Historical exploratory context only; raw data are not in this repository and no claim is current |
| Pogona TSD | mRNA | Planned only; requires a validated geometry/applicability assessment |
| Islet/diabetes multi-omic | mRNA + ATAC + genotype | Planned only; rank-deficient genotype layer is a required Stage 1 stress case |

---

## Stage 0 — Synthetic validation

The pipeline is tested against two synthetic ground truths.

### Two-group control

`synthetic_control()` generates two groups of samples with a planted signal
on the first component. Used to verify Stage 1 recovers the correct axis.

```{r synthetic-control}
library(landscapeR)

std <- synthetic_control(n = 40L, p = 500L, K = 2L, signal = 30, seed = 1L)
ctor <- get_strategy("Decomposer", "hogsvd_averaged")
std1 <- suppressWarnings(decompose(ctor(), std))@value

plot_spectrum(std1)
```

```{r component plots-synthetic}
plot_components(std1, colour_by = "group")
```

### Double-well potential control

`synthetic_potential_control()` draws samples from a known double-well
U(x) = (x² − 1)² via Langevin simulation. Provides a ground truth for
Stage 2 validation: wells at ±1, barrier at 0, barrier height = 1.

```{r double-well}
# synthetic_potential_control returns a 1-layer object (Langevin coordinates).
# Duplicate the layer so hogsvd_averaged (requires >= 2 layers) can run.
std_dw  <- synthetic_potential_control(n = 120L, seed = 42L)
se_dw   <- experiments(std_dw)[[1L]]
std_dw2l <- StateTransitionData(
  experiments = list(coords_a = se_dw, coords_b = se_dw),
  colData     = colData(std_dw),
  sampling_design = cross_sectional()
)

ctor    <- get_strategy("Decomposer", "hogsvd_averaged")
std_dw1 <- suppressWarnings(decompose(ctor(), std_dw2l))@value

dctor   <- get_strategy("DynamicsEstimator", "kde_logdensity")
std_dw2 <- estimate_dynamics(dctor(), std_dw1)@value
plot_potential(std_dw2)
```

> **Illustration only:** this single deterministic call exercises the estimator;
> it is not the multi-seed Stage 0 acceptance ladder or evidence for ADR 0002.

Recovery benchmark:

```{r benchmark}
# benchmark uses the original single-layer ground-truth object directly
bm <- potential_recovery_benchmark(std_dw)
knitr::kable(data.frame(
  metric = names(bm),
  value  = round(unlist(bm), 3)
), col.names = c("Metric", "Value"))
```

---

## Historical AML/CML exploratory context

Earlier exploratory work used AML/CML cohorts as a translation target for the
Rockne--Frankhouser literature. Those source data and local preprocessing
objects are deliberately **not** included in this public repository, so the
old AML chunks are removed rather than presented as runnable analysis.

They are also not evidence for the package's present claims. Under the
[evidence-first validation plan](https://github.com/drejom/landscapeR/blob/main/docs/implementation-plan-stage0-validation.md),
real-data Stage 2 inference is blocked until the frozen generic Stage 0 ladder
has passed, ADR 0002 has finalized thresholds, and a separate applicability and
confirmation workflow has been specified. Any future AML/CML article will
include a public data manifest, a fixed analysis specification, and the required
confirmation/claim-status diagnostics.

---

## Evidence status

The accepted Stage 1 v2 protocol compared two prototype estimators on
synthetic, heterogeneous-feature controls. Its completed, hash-verified full
artifact selected `C2_block_scaled_svd` in calibration, but C2 failed the
frozen holdout acceptance thresholds. Consequently no Stage 1 baseline is
accepted and #24 remains blocked. The [Stage 1 synthetic evidence article](stage1-evidence.html)
renders the immutable artifact; neither C1 substitution nor C2 tuning is
permitted from the consumed v2 holdout.

Similarly, the Stage 2 implementation is an auditable estimator, not yet a
validated biological inference engine. See ADR 0002 and the Stage 0 protocol
for its current claim boundary.

---

## Architecture notes

The package uses a **contract/registry pattern** — all stage algorithms are
S4 virtual classes with generics (`decompose`, `estimate_dynamics`). Concrete
strategies are registered by name and retrieved at runtime:

```{r registry}
list_strategies("Decomposer")
list_strategies("DynamicsEstimator")
```

This means adding a new algorithm (e.g. Kempf HO-GSVD, neural SDE) is a
`register_strategy()` call with no changes to calling code or tests.

---

*Last updated: 2026-07-13. Documentation site: <https://drejom.github.io/landscapeR/>.*
