---
title: "Development log"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Development log}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 8,
  fig.height = 5,
  dpi = 120
)
```

> **This is a living document.** It tracks what is working, what is being built,
> and what is planned. Updated after each development sprint.

---

## Roadmap

### Implemented

| Stage | Name | Status |
|---|---|---|
| Architecture | S4 contracts, registry, container, provenance | ✅ Done |
| 0 | Synthetic controls (two-group SVD, double-well Langevin) | ✅ Done |
| 1 | Comparative decomposition — `hogsvd_averaged`, `hogsvd_prereduced` | ✅ Done |
| 1+ | Multi-component output (k = 6), bimodality-sorted gallery | ✅ Done |
| 1+ | `project_into()` — Rockne2020 projection protocol | ✅ Done |
| 2 | KDE log-density inversion, constrained polynomial smooth | ✅ Done |

### Planned

| Stage | Name | Notes |
|---|---|---|
| 0.5 | Archetype classifier | `ranger`-based; assigns samples to known archetypes |
| 0.75 | Distributional fit assessment | SBI backend via `reticulate`; replaces ad-hoc BC threshold |
| 1 | HO-GSVD rank-deficiency handling | Kempf algorithm; critical for diabetes genotype layer |
| 2 | Minimum barrier height filter | Suppress noise wells in smooth unimodal KDE |
| 2 | Fokker-Planck time-to-transition | Solve FP equation; predict time to leukemia as in paper |
| 2 | Survival prediction | Validate against Kaplan-Meier; replicate paper Fig 4 |
| — | pkgdown site auto-deploy via CI | Render + rsync on push to main |
| — | CML reference replication | End-to-end on GSE244990; reproduce Leukemia 2024 Fig |

### Biological instantiations

| Dataset | Modality | Status |
|---|---|---|
| AML/CML (Cbfb-MYH11, BCR-ABL) | mRNA, miRNA | Data loaded, Stage 1 running |
| Pogona TSD | mRNA | Planned; sex-determination temperature axis |
| Islet/diabetes multi-omic | mRNA + ATAC + genotype | Planned; rank-deficient genotype layer |

---

## Stage 0 — Synthetic validation

The pipeline is tested against two synthetic ground truths.

### Two-group control

`synthetic_control()` generates two groups of samples with a planted signal
on the first component. Used to verify Stage 1 recovers the correct axis.

```{r synthetic-control}
library(landscapeR)

std <- synthetic_control(n = 40L, p = 500L, K = 2L, signal = 30, seed = 1L)
ctor <- get_strategy("Decomposer", "hogsvd_averaged")
std1 <- suppressWarnings(decompose(ctor(), std))@value

plot_spectrum(std1)
```

```{r gallery-synthetic}
plot_components(std1, colour_by = "group")
```

### Double-well potential control

`synthetic_potential_control()` draws samples from a known double-well
U(x) = (x² − 1)² via Langevin simulation. Provides a ground truth for
Stage 2 validation: wells at ±1, barrier at 0, barrier height = 1.

```{r double-well}
# synthetic_potential_control returns a 1-layer object (Langevin coordinates).
# Duplicate the layer so hogsvd_averaged (requires >= 2 layers) can run.
std_dw  <- synthetic_potential_control(n = 120L, seed = 42L)
se_dw   <- experiments(std_dw)[[1L]]
std_dw2l <- StateTransitionData(
  experiments = list(coords_a = se_dw, coords_b = se_dw),
  colData     = colData(std_dw)
)

ctor    <- get_strategy("Decomposer", "hogsvd_averaged")
std_dw1 <- suppressWarnings(decompose(ctor(), std_dw2l))@value

dctor   <- get_strategy("DynamicsEstimator", "kde_logdensity")
std_dw2 <- estimate_dynamics(dctor(), std_dw1)@value
plot_potential(std_dw2)
```

Recovery benchmark:

```{r benchmark}
# benchmark uses the original single-layer ground-truth object directly
bm <- potential_recovery_benchmark(std_dw)
knitr::kable(data.frame(
  metric = names(bm),
  value  = round(unlist(bm), 3)
), col.names = c("Metric", "Value"))
```

---

## Stage 1 — AML mRNA (primary 2018 cohort)

The 2018 cohort is the paper's primary experiment: clean randomised batches,
101 samples (T0–T6 + leukemic endpoint L). The 2016 cohort is batch-confounded
and treated as a supplementary stress-test, projected into the 2018 state-space.

```{r aml-stage1, eval=FALSE}
aml  <- readRDS("../data-raw/aml_mrna_std.rds")

# Run Stage 1 on the primary 2018 layer
ctor  <- get_strategy("Decomposer", "hogsvd_averaged")
std1  <- decompose(ctor(), aml)@value

# Gallery sorted by bimodality coefficient
# PC2 expected to be the AML state-transition axis (PC1 = age in paper)
plot_components(std1, colour_by = "condition")
```

> **Open question:** Gallery shows PC1 as highest BC (0.87) and η² (0.18) in
> the 2018 standalone SVD. The paper reports PC2 as the AML axis and PC1 as
> age. Verifying against paper figures — this section will be updated once
> confirmed.

---

## Stage 2 — Potential landscape

Stage 2 extracts the quasi-potential from the Stage 1 coordinates.
The target: recover the double-well landscape reported in Rockne2020,
with CTL samples in one well and CM/leukemic samples in the other.

```{r potential-aml, eval=FALSE}
dctor <- get_strategy("DynamicsEstimator", "kde_logdensity")
std2  <- estimate_dynamics(dctor(list(component = 2L)), std1)@value
plot_potential(std2)
```

> **Current status:** Stage 2 is implemented and runs. Over-fitting (too many
> wells/barriers) occurs when run on the wrong component (age axis, PC1).
> Running on PC2 pending confirmation of PC identity from paper.
> A minimum barrier height filter is planned for the next sprint.

---

## Architecture notes

The package uses a **contract/registry pattern** — all stage algorithms are
S4 virtual classes with generics (`decompose`, `estimate_dynamics`). Concrete
strategies are registered by name and retrieved at runtime:

```{r registry}
list_strategies("Decomposer")
list_strategies("DynamicsEstimator")
```

This means adding a new algorithm (e.g. Kempf HO-GSVD, neural SDE) is a
`register_strategy()` call with no changes to calling code or tests.

---

*Last updated: 2026-06-29*
