---
title: "Development log"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Development log}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 8,
  fig.height = 5,
  dpi = 120
)
library(landscapeR)
```

> **This is a living implementation record, not the package schedule or an
> acceptance-evidence report.** The root
> [ROADMAP](https://github.com/drejom/landscapeR/blob/main/ROADMAP.md) is the
> single scheduling authority. This article shows current user-visible behavior
> and links to implementation/evidence records. Acceptance evidence is published
> only from immutable, hash-verified benchmark artifacts.

---

## Current implementation status

| Stage | Name | Status |
|---|---|---|
| Architecture | S4 contracts, registry, container, provenance | ✅ Implemented and CI-tested |
| 0 | Synthetic controls (K=1/K≥2 subspace, exact-stationary double well) | ⚠️ Calibration controls implemented; frozen K=1 recovery/negative-control ladder remains #51 |
| 1 | Single-omic-layer registered SVD | ⚠️ Implemented and visually demonstrated; scientific acceptance remains blocked by #51 and #67 |
| 1 | Comparative decomposition — legacy equal-feature strategies | ⚠️ Implemented; not evidence for heterogeneous multi-omic fitting |
| 1 | Heterogeneous C1/C2 prototype and frozen comparison protocol | ⚠️ Full synthetic run complete; C2 won calibration but failed frozen holdout, so no baseline is accepted and #24 remains blocked |
| 1+ | Descriptive component gallery and legacy `project_into()` workflow | ⚠️ Gallery is implemented with canonical metadata alignment; projection remains blocked on #24 |
| 2 | KDE log-density inversion and constrained polynomial smooth | ⚠️ Implemented; ADR 0002 remains provisional-accepted |

Future sequencing, dependencies, and parked capabilities are maintained only in
[ROADMAP.md](https://github.com/drejom/landscapeR/blob/main/ROADMAP.md), so this
article does not duplicate a second planned-work list.

### Transient benchmark workspaces

Stage 1 benchmark helpers place their checkpoint workspaces under the
repository `.scratch/` directory by default. Set
`options(landscapeR.scratch_root = "/path/to/scratch")` before calling a
benchmark helper when transient state should live on another local or HPC
filesystem. An explicit `workspace` argument still takes precedence.

### Biological instantiations

| Dataset | Modality | Status |
|---|---|---|
| AML/CML (Cbfb-MYH11, BCR-ABL) | mRNA, miRNA | Historical exploratory context only; raw data are not in this repository and no claim is current |
| Pogona TSD | mRNA | Planned only; requires a validated geometry/applicability assessment |
| Islet/diabetes multi-omic | mRNA + ATAC + genotype | Planned only; rank-deficient genotype layer is a required Stage 1 stress case |

---

## Analysis intent lifecycle (AnalysisSpecification v2)

A component decision now **resolves rather than replaces** its target. Every
specification retains the target field and its neutral contrast/order/direction;
confirmation adds the frozen-basis component and analyst decision provenance.

```{r analysis-specification-v2}
proposal_id <- strrep("a", 64L)

spec_draft <- analysis_specification(
  id = "condition-axis",
  target_field = "condition",
  target_type = "binary",
  reference_level = "control",
  comparison_level = "disease"
)

spec_proposal <- analysis_specification(
  id = "condition-axis",
  target_field = "condition",
  target_type = "binary",
  reference_level = "control",
  comparison_level = "disease",
  lifecycle = "proposal",
  proposal_digest = proposal_id
)

spec_confirmed <- analysis_specification(
  id = "condition-axis",
  target_field = "condition",
  target_type = "binary",
  reference_level = "control",
  comparison_level = "disease",
  lifecycle = "confirmed",
  selected_component = 2L,
  proposal_digest = proposal_id,
  proposal_decision = "accepted",
  analyst_rationale = "Accepted the predeclared top-ranked target axis."
)

data.frame(
  lifecycle = c(spec_draft@lifecycle,
                spec_proposal@lifecycle,
                spec_confirmed@lifecycle),
  target = rep("control -> disease", 3L),
  proposal = c("absent", "present", "retained"),
  selected_component = c(NA_integer_, NA_integer_,
                         spec_confirmed@selected_component),
  decision = c("absent", "absent", spec_confirmed@proposal_decision)
)
```

Legacy v1 objects migrate only through `migrate_analysis_specification()`.
Target-only v1 objects require explicit direction; component-only v1 objects
also require the missing target and proposal decision record. Migration fails
rather than inventing intent or dropping the legacy component. See the
[complete v2 and migration contract](https://github.com/drejom/landscapeR/blob/main/docs/specs/analysis-specification.md).

---

## Stage 0 — Synthetic validation

The pipeline is tested against two synthetic ground truths.

Every successful Stage 0 generator records its complete generator parameters,
seed, and explicit claim status in provenance. Invalid potential controls raise
a typed validation error before they can become evidence. A `control_ladder()`
sweep retains failures as rows instead of discarding the rest of the grid. The
small proof below requests one supported two-layer cell and one unsupported
single-layer cell from the legacy HO-GSVD adapter.

```{r stage0-ladder-proof, echo=FALSE}
ladder_proof <- control_ladder(
  ns = 8L,
  ps = 8L,
  Ks = c(2L, 1L),
  signals = 20,
  seed = 122L
)
knitr::kable(
  ladder_proof[, c(
    "n", "p", "K", "signal", "status", "failure_class", "reason",
    "requested_cells", "completed_cells", "failed_cells"
  )],
  caption = paste(
    "Stage 0 control-ladder integrity proof. The supported cell completes;",
    "the unsupported cell retains its typed StageResult failure and reason."
  )
)
```

The table also carries requested, completed, and failed cell totals. This is an
implementation and calibration-input integrity contract, not evidence that a
decomposition strategy is scientifically accepted.

### Public validation boundary

Invalid declarations at public construction seams share one catchable package
condition. The matrix below is an implementation-safety proof: it records the
condition class and stable diagnostic available to a caller, without implying
scientific validation of any downstream analysis.

```{r typed-public-validation-proof}
validation_examples <- list(
  "Analysis specification" = function() analysis_specification(
    id = character(), target_field = "condition", target_type = "binary"
  ),
  "Publication theme" = function() theme_landscapeR(base_size = 0),
  "Observed-time storage" = function() {
    validation_data <- synthetic_control(
      n = 4L, p = 5L, K = 2L, signal = 10, seed = 124L
    )
    colData(validation_data)$collection_day <- c(
      "day 0", "day 1", "day 2", "day 3"
    )
    declare_sampling_design(
      validation_data,
      independent_time_course("collection_day")
    )
  }
)
validation_matrix <- do.call(rbind, lapply(
  names(validation_examples),
  function(boundary) {
    condition <- tryCatch(
      validation_examples[[boundary]](),
      error = identity
    )
    data.frame(
      `Public boundary` = boundary,
      `Condition class` = class(condition)[[1L]],
      Diagnostic = conditionMessage(condition),
      check.names = FALSE
    )
  }
))
knitr::kable(
  validation_matrix,
  caption = paste(
    "Public validation-boundary proof. Each invalid declaration is rejected",
    "as landscapeR_validation_error; the observed-time diagnostic names the",
    "affected metadata field before conversion is attempted."
  )
)
```

The table is a contract audit for software callers and contributors. It is not
an empirical performance result and does not establish biological validity.

### Two-group control

`synthetic_control()` generates two groups of samples with a planted signal
on the first component. Used to verify Stage 1 recovers the correct axis.

```{r synthetic-control}
library(landscapeR)

std <- synthetic_control(n = 40L, p = 500L, K = 2L, signal = 30, seed = 1L)
ctor <- get_strategy("Decomposer", "hogsvd_averaged")
std1 <- suppressWarnings(decompose(ctor(), std))@value

plot_spectrum(std1)
```

`hogsvd_averaged` and `hogsvd_prereduced` are legacy baseline adapters for
two or more layers defined on equal-sized feature spaces. They do not implement
the heterogeneous or rank-deficient HO-GSVD candidates being evaluated under
ADR 0001 and tracked in
[#49](https://github.com/drejom/landscapeR/issues/49). Their input boundary is
explicit:

| Input | Outcome |
|---|---|
| Two or more finite numeric layers with identical, unique, ordered feature identifiers | Typed Stage 1 success |
| Fewer than two layers | Typed failure with a reason |
| Different, missing, duplicated, or reordered feature identifiers | Typed unsupported-design failure with a reason |
| Invalid parameters, non-finite values, or numerical failure | Typed failure with a reason |

These outcomes describe implementation safety only. They do not constitute
scientific acceptance of either legacy strategy.

The component gallery aligns MAE-level metadata through each assay's canonical
sample map. Categorical metadata colour stored group-specific densities and
rugs; continuous metadata colour rugs beneath the stored overall density, so
rendering never refits a descriptive smoother. When a component or categorical
group has insufficient numerical spread for kernel-density estimation, its
sample coordinates remain visible as rugs and the separate scientific caption
identifies the unavailable density slice.

```{r component-gallery, fig.cap=c("Categorical descriptive gallery: planted synthetic groups colour densities and rugs after canonical sample alignment.", "Continuous descriptive gallery: the planted continuous coordinate colours the rug while the overall component density remains visible.")}
plot_components(
  std1,
  colour_by = "planted_group",
  n_components = 4L
)

plot_components(
  std1,
  colour_by = "u_shared",
  n_components = 4L
)
```

**Cold-reader conclusion:** metadata colour is visible for both categorical and
continuous fields, panels remain in decomposition order, and the gallery makes
no association score or component recommendation. Scientific ranking and analyst
confirmation remain the responsibility of the future atlas/proposal workflow.

### K=1 single-omic-layer SVD and double-well calibration (#50)

Before issue #50, the only runnable development-log path duplicated a
single-omic-layer coordinate assay so it could enter a K≥2 HO-GSVD strategy.
That workaround obscured the actual capability boundary and recorded the wrong
strategy for a K=1 analysis.

| Condition | Before #50 | After #50 |
|---|---|---|
| K=1 synthetic subspace | Rejected by the generator | First-class control with planted subspace truth |
| Stage 1 strategy | Duplicate the assay and call HO-GSVD | Registered exactly-one-omic-layer `svd` |
| Double-well observations | Thinned trajectory in the older estimator-level control | Independent exact-stationary draws for the K=1 end-to-end calibration control |
| Evidence status | Ambiguous development workaround | Explicitly `non_evidentiary_calibration` |

The current development path now keeps K=1 explicit from generation through
Stage 1 and the cross-sectional Stage 2 estimator:

```{r k1-double-well, fig.cap=c("Descriptive evidence: empirical density of the planted cross-sectional coordinate supplied to the single-omic-layer expression control.", "Hypothesis-conditioned interpretation: the fitted calibration-only quasi-potential; critical-point classification is deliberately omitted because uncertainty is not yet estimated.")}
std_k1 <- synthetic_k1_double_well_control(
  n = 200L, p = 100L, noise_sd = 0.03, seed = 50L
)

svd_ctor <- get_strategy("Decomposer", "svd")
std_k1_s1 <- suppressWarnings(decompose(svd_ctor(), std_k1))@value

coordinate_k1 <- colData(std_k1)$x_coord

empirical_density <- ggplot2::ggplot(
  data.frame(coordinate = coordinate_k1),
  ggplot2::aes(x = coordinate)
) +
  ggplot2::geom_density(fill = "#67A9CF", alpha = 0.45, linewidth = 0.8) +
  ggplot2::geom_rug(alpha = 0.35) +
  ggplot2::labs(
    title = "Descriptive evidence: planted cross-sectional coordinate density",
    x = "State-transition coordinate",
    y = "Density"
  ) +
  ggplot2::theme_bw(base_size = 11)
print(empirical_density)

dynamics_ctor <- get_strategy("DynamicsEstimator", "kde_logdensity")
std_k1_s2 <- estimate_dynamics(dynamics_ctor(), std_k1_s1)@value
plot_potential(std_k1_s2, show_critical_points = FALSE) +
  ggplot2::labs(
    subtitle = paste(
      "Critical-point classification omitted — uncertainty is not yet",
      "estimated; calibration-only output"
    )
  )
```

**Cold-reader conclusion:** the single-omic-layer expression control now passes
through the registered SVD strategy and produces the expected two-well
quasi-potential without fabricating a second omic layer. The public synthetic
input distribution remains visible before the fitted interpretation, and the
plot omits critical-point classification because uncertainty is not yet estimated.

This is implementation and calibration proof, not acceptance evidence. The
frozen K=1 thinness, negative-control, and independent acceptance protocol is
tracked in [#51](https://github.com/drejom/landscapeR/issues/51); the
AML-grounded target/confounder control is [#67](https://github.com/drejom/landscapeR/issues/67).

```{r k1-calibration-table}
k1_calibration <- suppressWarnings(k1_double_well_calibration(
  n = 200L, p = 100L, noise_sd = 0.03, seed = 50L
))

knitr::kable(data.frame(
  diagnostic = c(
    "Evidence status",
    "Stage 1 strategy",
    "Subspace angle (degrees)",
    "Mean well-location error",
    "Barrier-location error",
    "Barrier-height error"
  ),
  value = c(
    k1_calibration$evidence_status,
    k1_calibration$decomposer,
    format(round(k1_calibration$subspace_angle_deg, 3), nsmall = 3),
    format(round(k1_calibration$well_error, 3), nsmall = 3),
    format(round(k1_calibration$barrier_error, 3), nsmall = 3),
    format(round(k1_calibration$barrier_height_error, 3), nsmall = 3)
  )
), col.names = c("Diagnostic", "Observed calibration output"))
```

The table deliberately contains no pass, acceptance, or eligibility judgement.
Those decisions belong to the frozen independent protocol, not this disclosed
calibration run.

### Declared KDE bandwidth

Stage 2 treats bandwidth as part of the estimator specification rather than a
hidden fitting choice. The registered default remains `ks::hpi()`. A finite
positive numeric value can instead be declared for a Stage 0 sweep; the method
and resolved value are stored with the fitted result and its provenance.

```{r kde-bandwidth-proof, fig.width=4, fig.height=4, fig.cap="Default and explicitly declared bandwidths applied to the same frozen synthetic cross-sectional coordinate. Black shows the registered `ks::hpi()` estimator (resolved bandwidth printed in the legend); red shows the illustrative explicit value 0.25. Curves are shifted to a minimum of zero so their shapes can be compared without implying an absolute potential origin. This is implementation proof that bandwidth is sweepable and provenance-recorded; neither setting is accepted for biological inference from this visual comparison."}
default_kde <- estimate_dynamics(dynamics_ctor(), std_k1_s1)
explicit_kde <- estimate_dynamics(
  dynamics_ctor(params = list(
    bandwidth_method = "explicit",
    bandwidth_value = 0.25
  )),
  std_k1_s1
)

default_stage2 <- metadata(default_kde@value)$stage2
explicit_stage2 <- metadata(explicit_kde@value)$stage2
default_bandwidth_label <- sprintf(
  "Default hpi (h = %.3f)",
  default_stage2$bandwidth_value
)
bandwidth_curves <- rbind(
  data.frame(
    coordinate = default_stage2$x,
    quasi_potential = default_stage2$U - min(default_stage2$U),
    estimator = default_bandwidth_label
  ),
  data.frame(
    coordinate = explicit_stage2$x,
    quasi_potential = explicit_stage2$U - min(explicit_stage2$U),
    estimator = "Explicit (h = 0.250)"
  )
)

ggplot2::ggplot(
  bandwidth_curves,
  ggplot2::aes(x = coordinate, y = quasi_potential, colour = estimator)
) +
  ggplot2::geom_line(linewidth = 0.8) +
  scale_colour_landscapeR(
    "binary",
    reference_level = default_bandwidth_label,
    focal_level = "Explicit (h = 0.250)"
  ) +
  ggplot2::labs(
    x = "State-transition coordinate",
    y = "Relative quasi-potential",
    colour = "Bandwidth"
  ) +
  ggplot2::guides(colour = ggplot2::guide_legend(ncol = 1L)) +
  theme_landscapeR(base_size = 11) +
  ggplot2::theme(legend.position = "bottom")
```

```{r kde-typed-degenerate-proof}
degenerate_kde <- std_k1_s1
degenerate_stage1 <- metadata(degenerate_kde)$stage1
degenerate_stage1@coords_k <- lapply(
  degenerate_stage1@coords_k,
  function(coords) matrix(1, nrow = nrow(coords), ncol = ncol(coords))
)
degenerate_metadata <- metadata(degenerate_kde)
degenerate_metadata$stage1 <- degenerate_stage1
metadata(degenerate_kde) <- degenerate_metadata
degenerate_result <- estimate_dynamics(dynamics_ctor(), degenerate_kde)

knitr::kable(data.frame(
  outcome = degenerate_result@status,
  diagnostic = degenerate_result@reason
), col.names = c("Typed outcome", "Stable diagnostic"))
```

The degenerate coordinate is rejected before an external bandwidth or density
fit is attempted. Backend failures are likewise returned as typed Stage 2
failures, so a sweep can retain the failed cell without converting an external
`ks` exception into a pipeline crash.

---

## Historical AML/CML exploratory context

Earlier exploratory work used AML/CML cohorts as a translation target for the
Rockne--Frankhouser literature. The raw matrices and generated local
preprocessing objects are deliberately **not** included in this public
repository, so the old AML chunks are removed rather than presented as runnable
analysis. A minimal mapping of already-public mouse sample identifiers to the
checksum-pinned authoritative `sample_weeks` values is included solely to make
data lineage testable.

The corrected GSE133642 preparation keeps the source-study roles distinct from
landscapeR's use of both matrices as real biological validation data:

| Prepared layer | Source-study role | Observations | landscapeR role |
|---|---|---:|---|
| `mrna_primary_2018` | 2018 training cohort | 132 | future exploratory discovery basis |
| `mrna_supp_2016` | 2016 validation cohort 1 | 101 | later hostile batch/time-confounded projection test |

Its longitudinal declaration names only `mouse_id` and exact numeric
`sample_weeks` (unit: weeks). Categorical labels are inert, and no endpoint or
event semantics are inferred.

These data are not evidence for the package's present claims. Under the
[evidence-first validation plan](https://github.com/drejom/landscapeR/blob/main/docs/plans/implementation-plan-stage0-validation.md),
real-data Stage 2 inference is blocked until the frozen generic Stage 0 ladder
has passed, ADR 0002 has finalized thresholds, and a separate applicability and
confirmation workflow has been specified. Any future AML/CML article will
include a public data manifest, a fixed analysis specification, and the required
confirmation/claim-status diagnostics.

---

## Evidence status

The accepted Stage 1 v2 protocol compared two prototype estimators on
synthetic, heterogeneous-feature controls. Its completed, hash-verified full
artifact selected `C2_block_scaled_svd` in calibration, but C2 failed the
frozen holdout acceptance thresholds. Consequently no Stage 1 baseline is
accepted and #24 remains blocked. The [Stage 1 synthetic evidence article](stage1-evidence.html)
renders the immutable artifact; neither C1 substitution nor C2 tuning is
permitted from the consumed v2 holdout.

Similarly, the Stage 2 implementation is an auditable estimator, not yet a
validated biological inference engine. See ADR 0002 and the Stage 0 protocol
for its current claim boundary.

---

## Architecture notes

The package uses a **contract/registry pattern** — all stage algorithms are
S4 virtual classes with generics (`decompose`, `estimate_dynamics`). Concrete
strategies are registered by name and retrieved at runtime:

```{r registry}
list_strategies("Decomposer")
list_strategies("DynamicsEstimator")
```

This means adding a new algorithm (e.g. Kempf HO-GSVD, neural SDE) is a
`register_strategy()` call with no changes to calling code or tests.

---

## Implementation index

- **2026-07-14 — K=1 foundation and visual landing-proof policy:**
  [#50](https://github.com/drejom/landscapeR/issues/50) adds the registered
  single-omic-layer SVD and calibration control;
  [#68](https://github.com/drejom/landscapeR/issues/68) makes the pull request
  the canonical visual transition record. The full before/after proof is
  co-located with the diff in [PR #69](https://github.com/drejom/landscapeR/pull/69).
  This remains implementation work, not K=1 acceptance evidence.

*Last updated: 2026-07-14. Documentation site: <https://drejom.github.io/landscapeR/>.*
