---
title: "Development log"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Development log}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 8,
  fig.height = 5,
  dpi = 120
)
library(landscapeR)
```

> **This is a living document, not an acceptance-evidence report.** It tracks
> implementation progress and links to the governing protocols. Acceptance
> evidence is published only from immutable, hash-verified benchmark artifacts.

---

## Roadmap

### Implemented

| Stage | Name | Status |
|---|---|---|
| Architecture | S4 contracts, registry, container, provenance | ✅ Implemented and CI-tested |
| 0 | Synthetic controls (K=1/K≥2 subspace, exact-stationary double well) | ⚠️ Calibration controls implemented; frozen K=1 recovery/negative-control ladder remains #51 |
| 1 | Single-omic-layer registered SVD | ⚠️ Implemented and visually demonstrated; scientific acceptance remains blocked by #51 and #67 |
| 1 | Comparative decomposition — legacy equal-feature strategies | ⚠️ Implemented; not evidence for heterogeneous multi-omic fitting |
| 1 | Heterogeneous C1/C2 prototype and frozen comparison protocol | ⚠️ Full synthetic run complete; C2 won calibration but failed frozen holdout, so no baseline is accepted and #24 remains blocked |
| 1+ | Component gallery and legacy `project_into()` workflow | ⚠️ Exploratory/legacy; #24 is blocked on the selected heterogeneous strategy |
| 2 | KDE log-density inversion and constrained polynomial smooth | ⚠️ Implemented; ADR 0002 remains provisional-accepted |

### Planned

| Stage | Name | Notes |
|---|---|---|
| 0.5 | Archetype classifier | `ranger`-based; assigns samples to known archetypes |
| 0.75 | Distributional fit assessment | SBI backend via `reticulate`; replaces ad-hoc BC threshold |
| 1 | New heterogeneous-candidate decision | Requires a new ADR and frozen protocol; v2 holdout cannot be reused for tuning |
| 1 | HO-GSVD rank-deficiency handling | Kempf algorithm; candidate only after the current comparison decision |
| 2 | Full Stage 0 recovery and negative-control ladder | Establish ADR 0002 thresholds and pass rates |
| 2 | Minimum barrier height filter | Candidate only after Stage 0 evidence |
| 2 | Fokker-Planck time-to-transition | Requires a separately validated longitudinal strategy |
| 2 | Survival prediction | Requires validated real-data applicability and confirmation |
| — | Real-data confirmation workflows | Only after the generic Stage 0 and applicability gates pass |

### Biological instantiations

| Dataset | Modality | Status |
|---|---|---|
| AML/CML (Cbfb-MYH11, BCR-ABL) | mRNA, miRNA | Historical exploratory context only; raw data are not in this repository and no claim is current |
| Pogona TSD | mRNA | Planned only; requires a validated geometry/applicability assessment |
| Islet/diabetes multi-omic | mRNA + ATAC + genotype | Planned only; rank-deficient genotype layer is a required Stage 1 stress case |

---

## Stage 0 — Synthetic validation

The pipeline is tested against two synthetic ground truths.

### Two-group control

`synthetic_control()` generates two groups of samples with a planted signal
on the first component. Used to verify Stage 1 recovers the correct axis.

```{r synthetic-control}
library(landscapeR)

std <- synthetic_control(n = 40L, p = 500L, K = 2L, signal = 30, seed = 1L)
ctor <- get_strategy("Decomposer", "hogsvd_averaged")
std1 <- suppressWarnings(decompose(ctor(), std))@value

plot_spectrum(std1)
```

> The metadata-coloured component gallery is temporarily omitted here because
> its canonical MAE-level metadata alignment is being repaired in
> [#54](https://github.com/drejom/landscapeR/issues/54). The spectrum above is
> descriptive only and does not select a target biological axis.

### K=1 single-omic-layer SVD and double-well calibration (#50)

Before issue #50, the only runnable development-log path duplicated a
single-omic-layer coordinate assay so it could enter a K≥2 HO-GSVD strategy.
That workaround obscured the actual capability boundary and recorded the wrong
strategy for a K=1 analysis.

| Condition | Before #50 | After #50 |
|---|---|---|
| K=1 synthetic subspace | Rejected by the generator | First-class control with planted subspace truth |
| Stage 1 strategy | Duplicate the assay and call HO-GSVD | Registered exactly-one-omic-layer `svd` |
| Double-well observations | Thinned trajectory in the older estimator-level control | Independent exact-stationary draws for the K=1 end-to-end calibration control |
| Evidence status | Ambiguous development workaround | Explicitly `non_evidentiary_calibration` |

The current development path now keeps K=1 explicit from generation through
Stage 1 and the cross-sectional Stage 2 estimator:

```{r k1-double-well, fig.cap=c("Descriptive evidence: empirical density of the registered-SVD coordinate, shown before fitting a quasi-potential.", "Hypothesis-conditioned interpretation: the fitted calibration-only quasi-potential with stable and unstable critical points; this is not an acceptance result.")}
std_k1 <- synthetic_k1_double_well_control(
  n = 200L, p = 100L, noise_sd = 0.03, seed = 50L
)

svd_ctor <- get_strategy("Decomposer", "svd")
std_k1_s1 <- suppressWarnings(decompose(svd_ctor(), std_k1))@value

stage1_k1 <- metadata(std_k1_s1)$stage1
coordinate_k1 <- drop(dr_coords_k(stage1_k1)[[1L]][, 1L])

empirical_density <- ggplot2::ggplot(
  data.frame(coordinate = coordinate_k1),
  ggplot2::aes(x = coordinate)
) +
  ggplot2::geom_density(fill = "#67A9CF", alpha = 0.45, linewidth = 0.8) +
  ggplot2::geom_rug(alpha = 0.35) +
  ggplot2::labs(
    title = "Descriptive evidence: empirical SVD coordinate density",
    x = "State-transition coordinate",
    y = "Density"
  ) +
  ggplot2::theme_bw(base_size = 11)
print(empirical_density)

dynamics_ctor <- get_strategy("DynamicsEstimator", "kde_logdensity")
std_k1_s2 <- estimate_dynamics(dynamics_ctor(), std_k1_s1)@value
plot_potential(std_k1_s2)
```

**Cold-reader conclusion:** the single-omic-layer expression control now passes
through the registered SVD strategy and produces the expected two-well
quasi-potential without fabricating a second omic layer. The empirical density
and fitted quasi-potential remain visible together under ADR 0017.

This is implementation and calibration proof, not acceptance evidence. The
frozen K=1 thinness, negative-control, and independent acceptance protocol is
tracked in [#51](https://github.com/drejom/landscapeR/issues/51); the
AML-grounded target/confounder control is [#67](https://github.com/drejom/landscapeR/issues/67).

```{r k1-calibration-table}
k1_calibration <- suppressWarnings(k1_double_well_calibration(
  n = 200L, p = 100L, noise_sd = 0.03, seed = 50L
))

knitr::kable(data.frame(
  diagnostic = c(
    "Evidence status",
    "Stage 1 strategy",
    "Subspace angle (degrees)",
    "Mean well-location error",
    "Barrier-location error",
    "Barrier-height error"
  ),
  value = c(
    k1_calibration$evidence_status,
    k1_calibration$decomposer,
    format(round(k1_calibration$subspace_angle_deg, 3), nsmall = 3),
    format(round(k1_calibration$well_error, 3), nsmall = 3),
    format(round(k1_calibration$barrier_error, 3), nsmall = 3),
    format(round(k1_calibration$barrier_height_error, 3), nsmall = 3)
  )
), col.names = c("Diagnostic", "Observed calibration output"))
```

The table deliberately contains no pass, acceptance, or eligibility judgement.
Those decisions belong to the frozen independent protocol, not this disclosed
calibration run.

---

## Historical AML/CML exploratory context

Earlier exploratory work used AML/CML cohorts as a translation target for the
Rockne--Frankhouser literature. Those source data and local preprocessing
objects are deliberately **not** included in this public repository, so the
old AML chunks are removed rather than presented as runnable analysis.

They are also not evidence for the package's present claims. Under the
[evidence-first validation plan](https://github.com/drejom/landscapeR/blob/main/docs/implementation-plan-stage0-validation.md),
real-data Stage 2 inference is blocked until the frozen generic Stage 0 ladder
has passed, ADR 0002 has finalized thresholds, and a separate applicability and
confirmation workflow has been specified. Any future AML/CML article will
include a public data manifest, a fixed analysis specification, and the required
confirmation/claim-status diagnostics.

---

## Evidence status

The accepted Stage 1 v2 protocol compared two prototype estimators on
synthetic, heterogeneous-feature controls. Its completed, hash-verified full
artifact selected `C2_block_scaled_svd` in calibration, but C2 failed the
frozen holdout acceptance thresholds. Consequently no Stage 1 baseline is
accepted and #24 remains blocked. The [Stage 1 synthetic evidence article](stage1-evidence.html)
renders the immutable artifact; neither C1 substitution nor C2 tuning is
permitted from the consumed v2 holdout.

Similarly, the Stage 2 implementation is an auditable estimator, not yet a
validated biological inference engine. See ADR 0002 and the Stage 0 protocol
for its current claim boundary.

---

## Architecture notes

The package uses a **contract/registry pattern** — all stage algorithms are
S4 virtual classes with generics (`decompose`, `estimate_dynamics`). Concrete
strategies are registered by name and retrieved at runtime:

```{r registry}
list_strategies("Decomposer")
list_strategies("DynamicsEstimator")
```

This means adding a new algorithm (e.g. Kempf HO-GSVD, neural SDE) is a
`register_strategy()` call with no changes to calling code or tests.

---

## Implementation index

- **2026-07-14 — K=1 foundation and visual landing-proof policy:**
  [#50](https://github.com/drejom/landscapeR/issues/50) adds the registered
  single-omic-layer SVD and calibration control;
  [#68](https://github.com/drejom/landscapeR/issues/68) makes the pull request
  the canonical visual transition record. The full before/after proof is
  co-located with the diff in [PR #69](https://github.com/drejom/landscapeR/pull/69).
  This remains implementation work, not K=1 acceptance evidence.

*Last updated: 2026-07-14. Documentation site: <https://drejom.github.io/landscapeR/>.*
