---
title: "Map high-dimensional K=1 recovery"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Map high-dimensional K=1 recovery}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(landscapeR)
```

This calibration asks a narrow question: when does a planted biological
direction remain recoverable as the expression matrix becomes wider? It does
not estimate a universal sample-size requirement and it does not establish
biological validity.

Five regimes separate different reasons for increasing the number of features.
The fixed-total and fixed-sparse regimes add noise features without increasing
the planted signal. The coherent regime adds aligned informative features.
The correlated-module regime changes the noise covariance, and the null regime
checks that no target is recovered when no meaningful signal is present.

```{r workflow, eval=FALSE}
assessment <- run_k1_high_dimensional_calibration(
  feature_counts = c(100L, 1000L, 10000L),
  signal_ratios = c(0, 0.75, 1, 1.25),
  replicates = 20L,
  axis_resamples = 19L,
  seed = 42L
)

operating_map <- plot_k1_high_dimensional_calibration(assessment)
operating_map
scientific_caption(operating_map)
```

Panel A reports how often the known planted loading meets the declared cosine
threshold. Panel B reports how closely that axis recurs when biological
observations are resampled within target group. The pale vertical line is a
simulation reference: the independent-noise BBP value, adjusted by the largest
block-covariance eigenvalue for the correlated-module regime. It is not an
acceptance threshold.

The complete assessment retains the exact generator declaration, planted
answer, target recovery, typed association atlas and proposal or abstention,
axis-resampling evidence, downstream estimability, execution account, and RNG
streams. Large runs use the caller's future plan, or can be placed in a
`targets`/`crew` graph with `k1_high_dimensional_calibration_targets()`.
