---
title: "K=1 Stage 0 phase-B1 evidence"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{K=1 Stage 0 phase-B1 evidence}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 7, dpi = 120)
library(landscapeR)

artifact_name <- "k1-stage0-acceptance-v2-780f4b10ea21923a"
artifact_candidates <- c(
  file.path("inst", "benchmarks", artifact_name),
  file.path("..", "inst", "benchmarks", artifact_name),
  system.file("benchmarks", artifact_name, package = "landscapeR")
)
artifact_dir <- artifact_candidates[dir.exists(artifact_candidates)][1L]
stopifnot(length(artifact_dir) == 1L, !is.na(artifact_dir))
stopifnot(verify_k1_acceptance_artifact(artifact_dir))
summary <- readRDS(file.path(artifact_dir, "summary.rds"))
cells <- summary$cells

embed_artifact_figure <- function(filename, alt_text, caption_filename) {
  caption <- paste(
    readLines(file.path(artifact_dir, caption_filename), warn = FALSE),
    collapse = " "
  )
  cat(sprintf(
    '<figure><img src="%s" style="width:100%%" alt="%s"><figcaption>%s</figcaption></figure>',
    knitr::image_uri(file.path(artifact_dir, filename)),
    alt_text,
    caption
  ))
}
```

This article reports the frozen generic and negative-control phase of the K=1
Stage 0 acceptance experiment. It reads the committed, content-addressed
artifact and does not rerun simulations or change thresholds. The result is
negative but informative: landscapeR has not yet demonstrated a supported K=1
operating range under the complete frozen grid.

## What was tested

The experiment requested 16,100 independent tasks across double-well recovery,
pure-noise and single-well negative controls, and a shared-baseline missing-cell
safety design. It covered `n = 8, 12, 16, 24, 48, 96, 132, 192` independent
biological observations and `p = 100, 1,000, 10,000, 20,000` features. Every
requested task is retained in the artifact. Computational or model abstentions
remain in the denominator.

```{r execution-summary}
knitr::kable(data.frame(
  measure = c(
    "Requested tasks",
    "Successful result objects",
    "Complete execution",
    "Supported minimum n"
  ),
  value = c(
    summary$n_requested,
    summary$n_completed,
    summary$complete_execution,
    if (is.na(summary$supported_minimum_n)) "None" else summary$supported_minimum_n
  )
))
```

## Recovery and safety result

```{r pass-rate-figure, echo=FALSE, results="asis"}
embed_artifact_figure(
  "pass-rate.png",
  "K=1 independent acceptance pass rates across recovery and safety controls.",
  "pass-rate-caption.txt"
)
```

No double-well recovery cell passed. The best observed replicate pass rate was
78% at `n = 192`, `p = 100`, below both the 90% point-estimate gate and its
required 80% Wilson lower bound. Recovery was 0% at 10,000 and 20,000 features
for every tested sample size. All 100 shared-baseline replicates correctly
reported the interaction as non-estimable.

```{r false-positive-figure, echo=FALSE, results="asis"}
embed_artifact_figure(
  "false-positive.png",
  "K=1 negative-control false topology and target-selection rates.",
  "false-positive-caption.txt"
)
```

False target selection remained within the frozen limit: 0% in pure noise and
at most 2.5% in single-well data. False double-well topology did not. It reached
43.5% in pure noise and 45% in single-well data, so both negative controls fail
the prespecified 5% maximum.

## Interpretation and next step

The result does not justify a sample-size recommendation. Increasing sample
size improved some low-dimensional recovery cells, but did not make the method
safe across the frozen feature grid, and the topology estimator frequently
invented a second well in negative controls. These are algorithmic findings,
not evidence that more samples alone will solve the problem.

Issue #67 now runs the separately frozen AML-shaped acceptance control. Only
after that lane is incorporated can issue #51 make its final phase-B judgement
and amend ADR 0002. This article makes no biological-validity claim.

## Recovery provenance

Workers ran from reviewed merge
`7772baef1f7a69db107721487b20456a45f53bfe`. Collection exposed three contract
defects: vector iteration flattened typed target results, the validator
rejected legitimate missing errors when a well or barrier was not recovered,
and a sixteenth-decimal Wilson-bound difference prevented portable exact
verification. The 16,100 worker results were not changed or rerun. Tested
collection-only fixes preserved list results, accepted `NA` only for
non-estimable landmarks, and rounded Wilson bounds to 15 decimal places before
hashing. The artifact binds separate worker and collector identities, including
the governed recovery procedure and patch-file hashes. The recovery was
explicitly approved, and the permanent fixes are covered by regression tests.
