---
title: "Stage 1 heterogeneous synthetic evidence"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Stage 1 heterogeneous synthetic evidence}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 8, fig.height = 5, dpi = 120)
library(landscapeR)

artifact_name <- "stage1-heterogeneous-v2-a28239b9af0c5569e3be1892a5b60308c8451aefa04a165852cf521606087d4c"
artifact_candidates <- c(
  file.path("inst", "benchmarks", artifact_name),
  file.path("..", "inst", "benchmarks", artifact_name),
  system.file("benchmarks", artifact_name, package = "landscapeR")
)
artifact_dir <- artifact_candidates[dir.exists(artifact_candidates)][1L]
stopifnot(length(artifact_dir) == 1L, !is.na(artifact_dir))
evidence <- read_stage1_evidence_artifact(artifact_dir)
task_keys <- unique(evidence$results[c("stratum_digest", "seed")])
completed_tasks <- nrow(task_keys)
expected_tasks <- nrow(landscapeR:::.stage1_benchmark_strata(evidence$manifest)) * nrow(evidence$manifest$seeds)
expected_result_rows <- expected_tasks * length(evidence$manifest$candidates)
stopifnot(
  identical(completed_tasks, expected_tasks),
  identical(nrow(evidence$results), expected_result_rows),
  all(evidence$results$gate_passed)
)
```

This article renders the completed, frozen
[`stage1-heterogeneous-v2`](https://github.com/drejom/landscapeR/blob/main/docs/specs/stage1-candidate-protocol.md)
synthetic-evidence artifact. It does **not** rerun the benchmark, tune a
candidate, or make a biological claim. The complete artifact is hash-verified
before its contents are read.

## Result

The calibration rule selected `C2_block_scaled_svd`, but the frozen holdout
acceptance rule failed. Therefore **no heterogeneous Stage 1 baseline is
accepted** and Issue #24 remains blocked. C1 cannot be substituted after this
result: it was not the calibration-selected candidate, and the v2 holdout is
consumed.

```{r decision}
knitr::kable(data.frame(
  field = c("protocol", "protocol digest", "generator digest", "artifact verification",
            "completed tasks", "result rows", "source commit", "calibration selection",
            "holdout decision", "holdout thresholds passed"),
  value = c(
    evidence$manifest$protocol_id,
    evidence$selection$protocol_digest,
    evidence$selection$generator_digest,
    "complete payload and content-addressed name verified",
    completed_tasks,
    nrow(evidence$results),
    evidence$environment$commit,
    evidence$selection$selected_candidate,
    evidence$holdout$decision,
    evidence$holdout$thresholds_passed
  ),
  stringsAsFactors = FALSE
), col.names = c("Field", "Value"))
```

## Execution inventory

The verified table contains all 40,960 deterministic stratum-by-seed tasks and
no failed contract-gate rows. Expected missing-ID controls remain successful
when they emit their required typed failure.

```{r inventory}
knitr::kable(data.frame(
  field = c("completed deterministic tasks", "candidate-result rows", "failed contract-gate rows"),
  value = c(completed_tasks, nrow(evidence$results), sum(!evidence$results$gate_passed)),
  stringsAsFactors = FALSE
), col.names = c("Field", "Value"))
```

## Holdout threshold readout

The protocol requires every exact-ID stratum with shared signal 24 and noise
SD 1 to meet both thresholds. This table reports the worst median among those
frozen strata; a value above its threshold is sufficient to reject the
candidate.

```{r thresholds}
summary <- evidence$holdout$summary
required <- summary[
  summary$projection_case == "exact_ids" &
    summary$shared_signal == 24 &
    summary$noise_sd == 1 &
    summary$metric %in% c("shared_recovery_error", "projection_error"),
  , drop = FALSE
]
thresholds <- c(shared_recovery_error = 0.25, projection_error = 0.30)
threshold_readout <- do.call(rbind, lapply(names(thresholds), function(metric) {
  values <- required$estimate[required$metric == metric]
  data.frame(
    metric = metric,
    exact_id_strata = length(unique(required$stratum_digest[required$metric == metric])),
    worst_median = max(values),
    threshold = thresholds[[metric]],
    all_strata_pass = all(values <= thresholds[[metric]]),
    stringsAsFactors = FALSE
  )
}))
knitr::kable(transform(threshold_readout, worst_median = round(worst_median, 3)),
  col.names = c("Metric", "Exact-ID strata", "Worst median", "Threshold", "All pass"))
```

## Holdout figures

```{r figure-setup, include=FALSE}
# Parse stratum factor columns from the holdout summary string representation.
# The summary data.frame carries shared_signal and noise_sd directly.
# K is extracted from the stratum character string via gsub (preserves row
# alignment regardless of match success). missing_block is a boolean flag
# indicating whether any missing-block rate > 0 is present in the stratum.
exact <- as.data.frame(summary[
  summary$projection_case == "exact_ids" &
    summary$metric %in% c("shared_recovery_error", "projection_error"),
  , drop = FALSE
])
exact$K <- as.integer(gsub(".*K = ([0-9]+).*", "\\1", exact$stratum))
exact$missing_block <- grepl("missing_block_rate = 0\\.[1-9]", exact$stratum)
exact$signal_noise <- paste0(
  "signal ", exact$shared_signal, " / noise ", exact$noise_sd)
exact$col <- ifelse(exact$shared_signal == 24 & exact$noise_sd == 1,
  "#D73027",   # high signal, low noise (threshold-decisive strata)
  ifelse(exact$shared_signal == 24, "#FC8D59",  # high signal, high noise
    ifelse(exact$noise_sd == 1, "#91BFDB",      # low signal, low noise
      "#878787")))                               # low signal, high noise
exact$pch <- ifelse(exact$K == 2L, 16L, 17L)

# Helper: draw one metric panel
.draw_holdout_panel <- function(dat, metric_label, threshold, ylab_text) {
  if (nrow(dat) == 0L) {
    plot.new()
    title(main = metric_label, sub = "No data available")
    return(invisible(NULL))
  }
  idx <- seq_len(nrow(dat))
  plot(idx, dat$estimate,
    pch = dat$pch, col = dat$col,
    ylim = c(0, 1),
    xlab = "Canonical stratum index (sorted by grid order)",
    ylab = ylab_text,
    main = metric_label,
    las = 1L
  )
  has_ci <- !is.na(dat$ci_lower) & !is.na(dat$ci_upper) &
    dat$ci_lower != dat$ci_upper
  if (any(has_ci)) {
    arrows(idx[has_ci], dat$ci_lower[has_ci], idx[has_ci], dat$ci_upper[has_ci],
      angle = 90L, code = 3L, length = 0.02, col = dat$col[has_ci])
  }
  abline(h = threshold, lty = 2L, col = "black", lwd = 1.5)
  text(x = nrow(dat) * 0.95, y = threshold + 0.04,
    labels = paste("Acceptance threshold:", threshold),
    adj = 1L, cex = 0.8)
  legend("topleft",
    legend = c(
      "signal 24 / noise 1 (threshold-decisive)",
      "signal 24 / noise 2",
      "signal 12 / noise 1",
      "signal 12 / noise 2",
      "K = 2 (2-layer)",
      "K = 3 (3-layer)"
    ),
    pch = c(16L, 16L, 16L, 16L, 16L, 17L),
    col = c("#D73027", "#FC8D59", "#91BFDB", "#878787", "black", "black"),
    bty = "n", cex = 0.7, ncol = 2L
  )
  invisible(NULL)
}
```

```{r fig-shared-recovery, fig.width=9, fig.height=5.5, out.width="100%", fig.cap="**Figure 1. C2 holdout shared-subspace recovery error across all exact-ID strata.** Each point is the median shared-recovery error (lower is better) for one canonical stratum over 20 holdout seeds, with 95% bootstrap confidence intervals. Colour encodes the signal-to-noise regime: red = high signal (24) / low noise (1), the threshold-decisive conditions; orange = high signal / high noise; blue = low signal (12) / low noise; grey = low signal / high noise. Triangle symbols indicate 3-layer (K = 3) strata; circles indicate 2-layer (K = 2) strata. The dashed horizontal line is the frozen acceptance threshold (0.25). Points above the line are failing strata. The majority of strata fail even under high signal and low noise, indicating that C2 does not reliably recover the planted shared subspace in this heterogeneous-feature setting."}
sre_dat <- exact[exact$metric == "shared_recovery_error", ]
.draw_holdout_panel(sre_dat,
  metric_label = "Holdout shared-subspace recovery error (C2)",
  threshold = 0.25,
  ylab_text = "Shared recovery error (lower is better)"
)
```

```{r fig-projection, fig.width=9, fig.height=5.5, out.width="100%", fig.cap="**Figure 2. C2 holdout projection error across all exact-ID strata.** Each point is the median projection error (lower is better) for one canonical stratum over 20 holdout seeds, with 95% bootstrap confidence intervals. Colour and symbol encoding as in Figure 1. The dashed horizontal line is the frozen acceptance threshold (0.30). The low band of points (projection errors 0.22-0.35) corresponds to high-signal strata where the method partially captures the shared direction; the upper band (0.6-1.0) indicates near-complete failure. Even in the best-case high-signal, low-noise strata, a substantial fraction of replicates exceeds the threshold."}
pe_dat <- exact[exact$metric == "projection_error", ]
.draw_holdout_panel(pe_dat,
  metric_label = "Holdout projection error (C2)",
  threshold = 0.30,
  ylab_text = "Projection error (lower is better)"
)
```

## Provenance and next step

The checked artifact coverage comprises all 40,960 synthetic tasks and 81,920
candidate-result rows. Full artifacts are published only once every task
checkpoint is complete, so this result has no failed tasks. It also records the
seed manifest, protocol and generator digests, the calibration decision, and
the holdout report. Its source commit is
`6f1f0614b2c4f0d539baa69a20df1ef43705ade6`.

The negative confirmation result is retained as evidence. Any future attempt to
improve a candidate requires a new ADR and a newly frozen protocol; it must not
reuse or tune against this holdout.
