---
title: "Stage 1 heterogeneous synthetic evidence"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Stage 1 heterogeneous synthetic evidence}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 8, fig.height = 5, dpi = 120)
library(landscapeR)

artifact_name <- "stage1-heterogeneous-v2-a28239b9af0c5569e3be1892a5b60308c8451aefa04a165852cf521606087d4c"
artifact_candidates <- c(
  file.path("inst", "benchmarks", artifact_name),
  file.path("..", "inst", "benchmarks", artifact_name),
  system.file("benchmarks", artifact_name, package = "landscapeR")
)
artifact_dir <- artifact_candidates[dir.exists(artifact_candidates)][1L]
stopifnot(length(artifact_dir) == 1L, !is.na(artifact_dir))
evidence <- read_stage1_evidence_artifact(artifact_dir)
task_keys <- unique(evidence$results[c("stratum_digest", "seed")])
completed_tasks <- nrow(task_keys)
expected_tasks <- nrow(landscapeR:::.stage1_benchmark_strata(evidence$manifest)) * nrow(evidence$manifest$seeds)
expected_result_rows <- expected_tasks * length(evidence$manifest$candidates)
stopifnot(
  identical(completed_tasks, expected_tasks),
  identical(nrow(evidence$results), expected_result_rows),
  all(evidence$results$gate_passed)
)
```

This article renders the completed, frozen
[`stage1-heterogeneous-v2`](https://github.com/drejom/landscapeR/blob/main/docs/specs/stage1-candidate-protocol.md)
synthetic-evidence artifact. It does **not** rerun the benchmark, tune a
candidate, or make a biological claim. The complete artifact is hash-verified
before its contents are read.

## Production evidence orchestration

Production and scheduler-backed Stage 1 evidence runs use
`stage1_evidence_targets()`. The returned targets graph makes the frozen
manifest, task branches, complete-grid validation, calibration selection,
holdout assessment, atomic publication, and artifact verification separate
dependencies. `crew` controls where eligible targets run. Changing from a
local controller to PBS, Slurm, SGE, or LSF does not change a scientific
command, manifest identity, or RNG stream.
Scientific rows, selection, and holdout results are summarized by a dedicated
scientific digest that is invariant to the controller. The artifact content
address also covers execution provenance, so local and scheduler artifact
directory names may correctly differ even when their scientific digests match.

Each replicate branch independently checks the installed landscapeR revision,
R version, and package versions against the controller target before doing
scientific work. Worker results return to the controlling process. Publication
and hash verification run there only after every upstream target succeeds, so
an incomplete workflow cannot publish partial evidence as complete. A later
`tar_make()` retries failed or invalid branches from the durable targets store.

A local targets script can end with:

```r
controller <- stage1_crew_controller(
  scheduler = "local",
  name = "stage1-evidence",
  workers = 2L
)
targets::tar_option_set(controller = controller)

stage1_evidence_targets(
  artifact_root = "/absolute/path/to/evidence-artifacts",
  controller = "stage1-evidence"
)
```

Run that script with `targets::tar_make(use_crew = TRUE)`. Scheduler runs use
the same graph and replace only the controller. Scheduler resource options must
be supplied explicitly by the analysis owner. The legacy
`execute_stage1_benchmark_full()` function remains a local development and
migration adapter, not the production orchestration path.

## Summary-bootstrap execution

Current Stage 1 calibration and holdout summaries send both the paired-mean
bootstrap and the per-stratum median bootstrap through landscapeR's shared
future-backed repetition boundary. The package does not select a future plan:
users may choose sequential, multisession, cluster, or scheduler-backed
execution before calling the evidence functions. Each 10,000-repetition
summary records stable task identities, its declared run seed and derived
L'Ecuyer-CMRG streams, all requested/completed/failed counts, and serialized
input, output, provenance, and complete-execution measurements. Bootstrap draws
use the future seam's explicit legacy sequential-stream derivation: it captures
the original per-task RNG states in canonical order and performs each draw in
the worker, preserving the pre-migration 10,000-iteration estimates and
cross-metric pairing exactly.
Scheduling and worker-count defaults remain unchanged because the measured
tasks are small enough that local multisession startup and serialization can
dominate their runtime.
When an outer `targets` or `crew` task already owns parallel orchestration,
`sequential_internal = TRUE` executes repetition directly in the current
worker. It does not launch a nested future. Outside that boundary,
`future_scheduling` remains an explicit optional control rather than a package
default.

The committed v2 artifact below predates that execution-provenance extension
and is deliberately not regenerated or recalibrated. Its manifest, estimates,
thresholds, decision, and content address remain frozen. Newly executed
selection and holdout reports carry `bootstrap_executions` and
`bootstrap_measurements` alongside the same scientific summaries.

## Result

The calibration rule selected `C2_block_scaled_svd`, but the frozen holdout
acceptance rule failed. Therefore **no heterogeneous Stage 1 baseline is
accepted** and Issue #24 remains blocked. C1 cannot be substituted after this
result: it was not the calibration-selected candidate, and the v2 holdout is
consumed.

```{r decision}
knitr::kable(data.frame(
  field = c("protocol", "protocol digest", "generator digest", "artifact verification",
            "completed tasks", "result rows", "source commit", "calibration selection",
            "holdout decision", "holdout thresholds passed"),
  value = c(
    evidence$manifest$protocol_id,
    evidence$selection$protocol_digest,
    evidence$selection$generator_digest,
    "complete payload and content-addressed name verified",
    completed_tasks,
    nrow(evidence$results),
    evidence$environment$commit,
    evidence$selection$selected_candidate,
    evidence$holdout$decision,
    evidence$holdout$thresholds_passed
  ),
  stringsAsFactors = FALSE
), col.names = c("Field", "Value"))
```

## Execution inventory

The verified table contains all 40,960 deterministic stratum-by-seed tasks and
no failed contract-gate rows. Expected missing-ID controls remain successful
when they emit their required typed failure.

```{r inventory}
knitr::kable(data.frame(
  field = c("completed deterministic tasks", "candidate-result rows", "failed contract-gate rows"),
  value = c(completed_tasks, nrow(evidence$results), sum(!evidence$results$gate_passed)),
  stringsAsFactors = FALSE
), col.names = c("Field", "Value"))
```

## Holdout threshold readout

The protocol requires every exact-ID stratum with shared signal 24 and noise
SD 1 to meet both thresholds. This table reports the worst median among those
frozen strata; a value above its threshold is sufficient to reject the
candidate.

```{r thresholds}
summary <- evidence$holdout$summary
required <- summary[
  summary$projection_case == "exact_ids" &
    summary$shared_signal == 24 &
    summary$noise_sd == 1 &
    summary$metric %in% c("shared_recovery_error", "projection_error"),
  , drop = FALSE
]
thresholds <- c(shared_recovery_error = 0.25, projection_error = 0.30)
threshold_readout <- do.call(rbind, lapply(names(thresholds), function(metric) {
  values <- required$estimate[required$metric == metric]
  data.frame(
    metric = metric,
    exact_id_strata = length(unique(required$stratum_digest[required$metric == metric])),
    worst_median = max(values),
    threshold = thresholds[[metric]],
    all_strata_pass = all(values <= thresholds[[metric]]),
    stringsAsFactors = FALSE
  )
}))
knitr::kable(transform(threshold_readout, worst_median = round(worst_median, 3)),
  col.names = c("Metric", "Exact-ID strata", "Worst median", "Threshold", "All pass"))
```

## Holdout figures

```{r figure-setup, include=FALSE}
# Parse stratum factor columns from the holdout summary string representation.
# The summary data.frame carries shared_signal and noise_sd directly.
# K is extracted from the stratum character string via gsub (preserves row
# alignment regardless of match success). missing_block is a boolean flag
# indicating whether any missing-block rate > 0 is present in the stratum.
exact <- as.data.frame(summary[
  summary$projection_case == "exact_ids" &
    summary$metric %in% c("shared_recovery_error", "projection_error"),
  , drop = FALSE
])
exact$K <- as.integer(gsub(".*K = ([0-9]+).*", "\\1", exact$stratum))
exact$missing_block <- grepl("missing_block_rate = 0\\.[1-9]", exact$stratum)
exact$signal_noise <- paste0(
  "signal ", exact$shared_signal, " / noise ", exact$noise_sd)
exact$col <- ifelse(exact$shared_signal == 24 & exact$noise_sd == 1,
  "#D73027",   # high signal, low noise (threshold-decisive strata)
  ifelse(exact$shared_signal == 24, "#FC8D59",  # high signal, high noise
    ifelse(exact$noise_sd == 1, "#91BFDB",      # low signal, low noise
      "#878787")))                               # low signal, high noise
exact$pch <- ifelse(exact$K == 2L, 16L, 17L)

# Helper: draw one metric panel
.draw_holdout_panel <- function(dat, metric_label, threshold, ylab_text) {
  if (nrow(dat) == 0L) {
    plot.new()
    title(main = metric_label, sub = "No data available")
    return(invisible(NULL))
  }
  idx <- seq_len(nrow(dat))
  plot(idx, dat$estimate,
    pch = dat$pch, col = dat$col,
    ylim = c(0, 1),
    xlab = "Canonical stratum index (sorted by grid order)",
    ylab = ylab_text,
    main = metric_label,
    las = 1L
  )
  has_ci <- !is.na(dat$ci_lower) & !is.na(dat$ci_upper) &
    dat$ci_lower != dat$ci_upper
  if (any(has_ci)) {
    arrows(idx[has_ci], dat$ci_lower[has_ci], idx[has_ci], dat$ci_upper[has_ci],
      angle = 90L, code = 3L, length = 0.02, col = dat$col[has_ci])
  }
  abline(h = threshold, lty = 2L, col = "black", lwd = 1.5)
  text(x = nrow(dat) * 0.95, y = threshold + 0.04,
    labels = paste("Acceptance threshold:", threshold),
    adj = 1L, cex = 0.8)
  legend("topleft",
    legend = c(
      "signal 24 / noise 1 (threshold-decisive)",
      "signal 24 / noise 2",
      "signal 12 / noise 1",
      "signal 12 / noise 2",
      "K = 2 (2-layer)",
      "K = 3 (3-layer)"
    ),
    pch = c(16L, 16L, 16L, 16L, 16L, 17L),
    col = c("#D73027", "#FC8D59", "#91BFDB", "#878787", "black", "black"),
    bty = "n", cex = 0.7, ncol = 2L
  )
  invisible(NULL)
}
```

```{r fig-shared-recovery, fig.width=9, fig.height=5.5, out.width="100%", fig.cap="**Figure 1. C2 holdout shared-subspace recovery error across all exact-ID strata.** Each point is the median shared-recovery error (lower is better) for one canonical stratum over 20 holdout seeds, with 95% bootstrap confidence intervals. Colour encodes the signal-to-noise regime: red = high signal (24) / low noise (1), the threshold-decisive conditions; orange = high signal / high noise; blue = low signal (12) / low noise; grey = low signal / high noise. Triangle symbols indicate 3-layer (K = 3) strata; circles indicate 2-layer (K = 2) strata. The dashed horizontal line is the frozen acceptance threshold (0.25). Points above the line are failing strata. The majority of strata fail even under high signal and low noise, indicating that C2 does not reliably recover the planted shared subspace in this heterogeneous-feature setting."}
sre_dat <- exact[exact$metric == "shared_recovery_error", ]
.draw_holdout_panel(sre_dat,
  metric_label = "Holdout shared-subspace recovery error (C2)",
  threshold = 0.25,
  ylab_text = "Shared recovery error (lower is better)"
)
```

```{r fig-projection, fig.width=9, fig.height=5.5, out.width="100%", fig.cap="**Figure 2. C2 holdout projection error across all exact-ID strata.** Each point is the median projection error (lower is better) for one canonical stratum over 20 holdout seeds, with 95% bootstrap confidence intervals. Colour and symbol encoding as in Figure 1. The dashed horizontal line is the frozen acceptance threshold (0.30). The low band of points (projection errors 0.22-0.35) corresponds to high-signal strata where the method partially captures the shared direction; the upper band (0.6-1.0) indicates near-complete failure. Even in the best-case high-signal, low-noise strata, a substantial fraction of replicates exceeds the threshold."}
pe_dat <- exact[exact$metric == "projection_error", ]
.draw_holdout_panel(pe_dat,
  metric_label = "Holdout projection error (C2)",
  threshold = 0.30,
  ylab_text = "Projection error (lower is better)"
)
```

## Provenance and next step

The checked artifact coverage comprises all 40,960 synthetic tasks and 81,920
candidate-result rows. Full artifacts are published only once every task
checkpoint is complete, so this result has no failed tasks. It also records the
seed manifest, protocol and generator digests, the calibration decision, and
the holdout report. Its source commit is
`6f1f0614b2c4f0d539baa69a20df1ef43705ade6`.

The negative confirmation result is retained as evidence. Any future attempt to
improve a candidate requires a new ADR and a newly frozen protocol; it must not
reuse or tune against this holdout.
