---
title: "Stage 1 heterogeneous synthetic evidence"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Stage 1 heterogeneous synthetic evidence}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 8, fig.height = 5, dpi = 120)
library(landscapeR)

artifact_name <- "stage1-heterogeneous-v2-a28239b9af0c5569e3be1892a5b60308c8451aefa04a165852cf521606087d4c"
artifact_candidates <- c(
  file.path("inst", "benchmarks", artifact_name),
  file.path("..", "inst", "benchmarks", artifact_name),
  system.file("benchmarks", artifact_name, package = "landscapeR")
)
artifact_dir <- artifact_candidates[dir.exists(artifact_candidates)][1L]
stopifnot(length(artifact_dir) == 1L, !is.na(artifact_dir))
evidence <- read_stage1_evidence_artifact(artifact_dir)
task_keys <- unique(evidence$results[c("stratum_digest", "seed")])
completed_tasks <- nrow(task_keys)
expected_tasks <- nrow(landscapeR:::.stage1_benchmark_strata(evidence$manifest)) * nrow(evidence$manifest$seeds)
expected_result_rows <- expected_tasks * length(evidence$manifest$candidates)
stopifnot(
  identical(completed_tasks, expected_tasks),
  identical(nrow(evidence$results), expected_result_rows),
  all(evidence$results$gate_passed)
)
```

This article renders the completed, frozen
[`stage1-heterogeneous-v2`](https://github.com/drejom/landscapeR/blob/main/docs/specs/stage1-candidate-protocol.md)
synthetic-evidence artifact. It does **not** rerun the benchmark, tune a
candidate, or make a biological claim. The complete artifact is hash-verified
before its contents are read.

## Result

The calibration rule selected `C2_block_scaled_svd`, but the frozen holdout
acceptance rule failed. Therefore **no heterogeneous Stage 1 baseline is
accepted** and Issue #24 remains blocked. C1 cannot be substituted after this
result: it was not the calibration-selected candidate, and the v2 holdout is
consumed.

```{r decision}
knitr::kable(data.frame(
  field = c("protocol", "protocol digest", "generator digest", "artifact verification",
            "completed tasks", "result rows", "source commit", "calibration selection",
            "holdout decision", "holdout thresholds passed"),
  value = c(
    evidence$manifest$protocol_id,
    evidence$selection$protocol_digest,
    evidence$selection$generator_digest,
    "complete payload and content-addressed name verified",
    completed_tasks,
    nrow(evidence$results),
    evidence$environment$commit,
    evidence$selection$selected_candidate,
    evidence$holdout$decision,
    evidence$holdout$thresholds_passed
  ),
  stringsAsFactors = FALSE
), col.names = c("Field", "Value"))
```

## Execution inventory

The verified table contains all 40,960 deterministic stratum-by-seed tasks and
no failed contract-gate rows. Expected missing-ID controls remain successful
when they emit their required typed failure.

```{r inventory}
knitr::kable(data.frame(
  field = c("completed deterministic tasks", "candidate-result rows", "failed contract-gate rows"),
  value = c(completed_tasks, nrow(evidence$results), sum(!evidence$results$gate_passed)),
  stringsAsFactors = FALSE
), col.names = c("Field", "Value"))
```

## Holdout threshold readout

The protocol requires every exact-ID stratum with shared signal 24 and noise
SD 1 to meet both thresholds. This table reports the worst median among those
frozen strata; a value above its threshold is sufficient to reject the
candidate.

```{r thresholds}
summary <- evidence$holdout$summary
required <- summary[
  summary$projection_case == "exact_ids" &
    summary$shared_signal == 24 &
    summary$noise_sd == 1 &
    summary$metric %in% c("shared_recovery_error", "projection_error"),
  , drop = FALSE
]
thresholds <- c(shared_recovery_error = 0.25, projection_error = 0.30)
threshold_readout <- do.call(rbind, lapply(names(thresholds), function(metric) {
  values <- required$estimate[required$metric == metric]
  data.frame(
    metric = metric,
    exact_id_strata = length(unique(required$stratum_digest[required$metric == metric])),
    worst_median = max(values),
    threshold = thresholds[[metric]],
    all_strata_pass = all(values <= thresholds[[metric]]),
    stringsAsFactors = FALSE
  )
}))
knitr::kable(transform(threshold_readout, worst_median = round(worst_median, 3)),
  col.names = c("Metric", "Exact-ID strata", "Worst median", "Threshold", "All pass"))
```

## Frozen figures

```{r figures, fig.cap=c("Exact-ID holdout shared-recovery error by canonical stratum.", "Exact-ID holdout projection error by canonical stratum."), out.width="100%"}
exact <- summary[
  summary$projection_case == "exact_ids" &
    summary$metric %in% c("shared_recovery_error", "projection_error"),
  , drop = FALSE
]
for (metric in c("shared_recovery_error", "projection_error")) {
  figure_data <- exact[exact$metric == metric, , drop = FALSE]
  plot(
    seq_len(nrow(figure_data)), figure_data$estimate,
    pch = 16L, col = "#2C7FB8",
    xlab = "Canonical holdout stratum", ylab = metric,
    main = paste("Stage 1 v2 holdout", metric)
  )
  nonzero_ci <- figure_data$ci_lower != figure_data$ci_upper
  if (any(nonzero_ci)) {
    arrows(
      which(nonzero_ci), figure_data$ci_lower[nonzero_ci],
      which(nonzero_ci), figure_data$ci_upper[nonzero_ci],
      angle = 90L, code = 3L, length = 0.03
    )
  }
}
```

## Provenance and next step

The checked artifact coverage comprises all 40,960 synthetic tasks and 81,920
candidate-result rows. Full artifacts are published only once every task
checkpoint is complete, so this result has no failed tasks. It also records the
seed manifest, protocol and generator digests, the calibration decision, and
the holdout report. Its source commit is
`6f1f0614b2c4f0d539baa69a20df1ef43705ade6`.

The negative confirmation result is retained as evidence. Any future attempt to
improve a candidate requires a new ADR and a newly frozen protocol; it must not
reuse or tune against this holdout.
